---
title: "CompoundDb"
author: "Mar Garcia-Aloy"
output: 
  BiocStyle::html_document:
    toc: false
    number_sections: false
    toc_float: false
---

https://rformassspectrometry.github.io/CompoundDb/articles/create-compounddb.html#compdb-from-custom-data

```{r, include = FALSE}
startpoint <- Sys.time()
```

# Libraries

```{r, message = FALSE}
library(tidyverse)
library(Rdisop)
library(CompoundDb)
library(MetaboCoreUtils)
```


# Main code

Define a function that allows to calculate molecular formulas according to 
each lipid class structure/backbone and the total number of carbons and 
double bonds of their fatty acyl chains.

```{r}
fml_maker <- function(class, C, db){
  case_when(
    # 1 FA chain -----------------------------------------------------------
    class == "FFA" ~ paste0("C", C, "H", C*2 - (2*db), "O2"),
    class == "CAR" ~ paste0("C", C + 7, "H", C*2 - (2*db) + 13, "NO4"),
    class == "LPC" ~ paste0("C", C + 8, "H", C*2 - (2*db) + 18, "NO7P"),
    class == "LPE" ~ paste0("C", C + 5, "H", C*2 - (2*db) + 12, "NO7P"),
    class == "LPI" ~ paste0("C", C + 9, "H", C*2 - (2*db) + 17, "O12P"),
    class == "DGMG" ~ paste0("C", C + 15, "H", C*2 - (2*db) + 26, "O14"),
    class == "MG"   ~ paste0("C", C +  3, "H", C*2 - (2*db) +  6, "O4"),
    class == "ST" ~ paste0("C", C +  29, "H", C*2 - (2*db) + 48, "O2"),
    class == "Glc-ST" ~ paste0("C", C +  35, "H", C*2 - (2*db) + 58, "O7"),
    
    # 2 FA chain -----------------------------------------------------------
    class == "PA"  ~ paste0("C", C + 3, "H", C*2 - (2*db) +  5, "O8P"),
    class == "mPA" ~ paste0("C", C + 4, "H", C*2 - (2*db) +  7, "O8P"),
    class == "dmPA"~ paste0("C", C + 5, "H", C*2 - (2*db) +  9, "O8P"),
    class == "PC"  ~ paste0("C", C + 8, "H", C*2 - (2*db) + 16, "NO8P"),
    class == "PE"  ~ paste0("C", C + 5, "H", C*2 - (2*db) + 10, "NO8P"),
    class == "PG"  ~ paste0("C", C + 6, "H", C*2 - (2*db) + 11, "O10P"),
    class == "PI"  ~ paste0("C", C + 9, "H", C*2 - (2*db) + 15, "O13P"),
    class == "PS"  ~ paste0("C", C + 6, "H", C*2 - (2*db) + 10, "NO10P"),
    
    class == "SM"       ~ paste0("C", C + 4, "H", C*2 - (2*db) + 11,"N2O6P"),
    class == "Cer"      ~ paste0("C", C,     "H", C*2 - (2*db) +  1, "NO3"),
    class == "Cer;O3"   ~ paste0("C", C,     "H", C*2 - (2*db) +  1, "NO4"),
    class == "Cer;O4"   ~ paste0("C", C,     "H", C*2 - (2*db) +  1, "NO5"),
    class == "HexCer"   ~ paste0("C", C + 6, "H", C*2 - (2*db) + 11, "NO8"),
    class == "HexCer;O3"~ paste0("C", C + 6, "H", C*2 - (2*db) + 11, "NO9"),
    class == "HexCer;O4"~ paste0("C", C + 6, "H", C*2 - (2*db) + 11, "NO10"),
    class == "LactCer"  ~ paste0("C", C +12, "H", C*2 - (2*db) + 21, "NO13"),
    
    class == "MGDG" ~ paste0("C", C +  9, "H", C*2 - (2*db) + 14, "O10"),
    class == "DGDG" ~ paste0("C", C + 15, "H", C*2 - (2*db) + 24, "O15"),
    class == "SQDG" ~ paste0("C", C +  9, "H", C*2 - (2*db) + 14, "O12S"),
    class == "acMGDG"~paste0("C", C +  9, "H", C*2 - (2*db) + 12, "O11"),
    
    class == "DG"   ~ paste0("C", C +  3, "H", C*2 - (2*db) +  4, "O5"),
    
    # 3 FA chain -----------------------------------------------------------
    class == "TG"   ~ paste0("C", C +  3, "H", C*2 - (2*db) +  2, "O6")
  )
}
```

Generate all potential lipid species build with fatty acyl chains with a 
number of carbons between 12 and 24, and with up to 6 unsaturations, except 
for sphingolipids for which the potential number of carbons of fatty acyl 
chains is extended until 28.

```{r}
cmps <- c()

cls <- c("FFA", "CAR", "LPC", "LPE", "LPI", "MG", "DGMG", "ST", "Glc-ST")
C <- seq(from = 12, to = 24, by = 1)
db <- seq(from = 0, to = 6, by = 1)
sn <-  paste(expand.grid(C, db)[,"Var1"], expand.grid(C, db)[,"Var2"], 
             sep = ":")
for(i in cls){
  cmps <- c(cmps, paste(i, sn))
}

cls <- c("PA", "mPA", "dmPA", "PC", "PE", "PG", "PI", "PS", 
         "MGDG", "DGDG", "SQDG", "DG")
C <- seq(from = 12*2, to = 24*2, by = 1)
db <- seq(from = 0, to = 6*2, by = 1)
sn <-  paste(expand.grid(C, db)[,"Var1"], expand.grid(C, db)[,"Var2"], 
             sep = ":")
for(i in cls){
  cmps <- c(cmps, paste(i, sn))
}

cls <- c("SM", "Cer", "Cer;O3", "Cer;O4", 
         "HexCer", "HexCer;O3", "HexCer;O4", "LactCer")
C <- seq(from = 12*2, to = 28*2, by = 1)
db <- seq(from = 0, to = 6*2, by = 1)
sn <-  paste(expand.grid(C, db)[,"Var1"], expand.grid(C, db)[,"Var2"], 
             sep = ":")
for(i in cls){
  cmps <- c(cmps, paste(i, sn))
}

cls <- c("TG", "acMGDG")
C <- seq(from = 12*3, to = 24*3, by = 1)
db <- seq(from = 0, to = 6*3, by = 1)
sn <-  paste(expand.grid(C, db)[,"Var1"], expand.grid(C, db)[,"Var2"], 
             sep = ":")
for(i in cls){
  cmps <- c(cmps, paste(i, sn))
}

rm(cls, C, db, sn, i)
```


Create a `CompDb` database from automatically defined compounds.  
Create a `data.frame` with compound information of candidate lipids.   
The data needs to have the following information "compound_id", "name", 
"inchi", "inchikey", "formula", "exactmass", "synonyms".  
All columns except "compound_id" can also contain missing values.    
It is also possible to define additional columns.  

```{r}
cmps <- data.frame(
  compound_id = paste0(
    "C", formatC(seq(length(cmps)), width = nchar(length(cmps)), flag = "0")),
  name = cmps,
  inchi = NA_character_,
  inchikey = NA_character_,
  formula = NA,
  exactmass = NA,
  synonyms = NA_character_,
  class = gsub("\\ .*", "", cmps),
  adduct = c("[M+H]+ ; [2M+H]+ ; [M-H]- ; [2M-H]-"), 
  rtime = NA,
  std = FALSE,
  instd = FALSE,
  C = gsub(":.*", "", gsub(".*\\ ", "", cmps)),
  db = gsub(".*:", "", gsub(".*\\ ", "", cmps)),
  t_min = NA,
  t_max = NA,
  m_min = NA,
  m_max = NA
)
cmps$C <- as.numeric(cmps$C)
cmps$db <- as.numeric(cmps$db)
for(i in seq(nrow(cmps))){
  cmps$formula[i] <- fml_maker(cmps$class[i], cmps$C[i], cmps$db[i])
}
rm(i)
```

Add manual curated data:

```{r}
cmps_RT <- read.csv("data/compounds.csv")
cmps_RT <- cmps_RT[!cmps_RT$compound %in% c(
  "PA 18:1_18:3", "PC 18:1_18:3",  "PE 18:1_18:3"#, #"PE 18:1_18:1","PG 18:1_18:3"
),]
cmps_RT$name <- paste0(cmps_RT$class, " ", cmps_RT$C, ":", cmps_RT$db)
cmps_RT_IS <- cmps_RT[cmps_RT$IS == TRUE, ]
cmps_RT_others <- cmps_RT[cmps_RT$IS == FALSE, ]
cmps_RT_others <- cmps_RT_others[!cmps_RT_others$class %in% cmps$class,]
cmps_RT <- cmps_RT[cmps_RT$IS == FALSE, ]
cmps_RT <- cmps_RT[cmps_RT$class %in% cmps$class,]
for(i in seq(nrow(cmps))){
  idx <- which(cmps$name[i] == cmps_RT$name)
  if(length(idx) > 1){
    cmps <- rbind(
      cmps, 
      do.call("rbind", 
              replicate(length(idx)-1, cmps[i,], simplify = FALSE)))
    cmps$compound_id[(nrow(cmps) - (length(idx)-2)):nrow(cmps)] <- paste0(
      "C", as.numeric(gsub(
        "C", "", 
        cmps$compound_id[nrow(cmps)-(length(idx)-1)])) + seq(length(idx)-1))
    idx2 <- c(i, (nrow(cmps) - (length(idx)-2)):nrow(cmps))
    cmps$name[idx2] <- cmps_RT$compound[idx]
    cmps$rtime[idx2] <- cmps_RT$rtime[idx]
    cmps$std[idx2] <- cmps_RT$standard[idx]
    cmps$t_min[idx2] <- cmps_RT$t_min[idx]
    cmps$t_max[idx2] <- cmps_RT$t_max[idx]
    cmps$m_min[idx2] <- cmps_RT$m_min[idx]
    cmps$m_max[idx2] <- cmps_RT$m_max[idx]
  } else if(length(idx) == 1){
    cmps$name[i] <- cmps_RT$compound[idx]
    cmps$rtime[i] <- cmps_RT$rtime[idx]
    cmps$std[i] <- cmps_RT$standard[idx]
    cmps$t_min[i] <- cmps_RT$t_min[idx]
    cmps$t_max[i] <- cmps_RT$t_max[idx]
    cmps$m_min[i] <- cmps_RT$m_min[idx]
    cmps$m_max[i] <- cmps_RT$m_max[idx]
  }
}
cmps <- rbind(
  cmps, 
  data.frame(
    compound_id = paste0(
      "C", as.numeric(gsub("C", "", cmps$compound_id[nrow(cmps)])
      ) + seq(nrow(cmps_RT_IS))),
    name = cmps_RT_IS$compound,
    inchi = NA_character_,
    inchikey = NA_character_,
    formula = cmps_RT_IS$formula,
    exactmass = NA,
    synonyms = cmps_RT_IS$name,
    class = cmps_RT_IS$class,
    adduct = rep(c("[M+H]+ ; [2M+H]+ ; [M-H]- ; [2M-H]-"), nrow(cmps_RT_IS)),
    rtime = cmps_RT_IS$rtime,
    std = TRUE,
    instd = TRUE,
    C = cmps_RT_IS$C,
    db = cmps_RT_IS$db,
    t_min = cmps_RT_IS$t_min,
    t_max = cmps_RT_IS$t_max,
    m_min = cmps_RT_IS$m_min,
    m_max = cmps_RT_IS$m_max
  ))
cmps <- rbind(
  cmps, 
  data.frame(
    compound_id = paste0(
      "C", as.numeric(gsub(
        "C", "", cmps$compound_id[nrow(cmps)])) + seq(nrow(cmps_RT_others))),
    name = cmps_RT_others$compound,
    inchi = NA_character_,
    inchikey = NA_character_,
    formula = cmps_RT_others$formula,
    exactmass = NA,
    synonyms = cmps_RT_others$name,
    class = cmps_RT_others$class,
    adduct = cmps_RT_others$comments,
    rtime = cmps_RT_others$rtime,
    std = FALSE,
    instd = FALSE,
    C = cmps_RT_others$C,
    db = cmps_RT_others$db,
    t_min = cmps_RT_others$t_min,
    t_max = cmps_RT_others$t_max,
    m_min = cmps_RT_others$m_min,
    m_max = cmps_RT_others$m_max
  ))
for(i in seq(nrow(cmps))){
  if(cmps$instd[i] == "TRUE"){
    cmps$exactmass[i] <- getMolecule(cmps$formula[i])$exactmass
  } else {
    cmps$exactmass[i] <- calculateMass(cmps$formula[i])
  }
  if(cmps$class[i] %in% c("FFA", "LPA", "LPG", "LPI", 
                          "PA", "mPA", "dmPA", "PG", "PI", "SQDG", 
                          "MG", "DG", "TG")){
    cmps$adduct[i] <- "[M+NH4]+ ; [2M+NH4]+ ; [M-H]- ; [2M-H]-"
  } else if(cmps$class[i] %in% c("MGDG", "DGDG", "acMGDG", "ST", "Glc-ST")){
    cmps$adduct[i] <- "[M+NH4]+ ; [2M+NH4]+ ; [M+CHO2]- ; [2M+CHO2]-"
  } else if(cmps$class[i] %in% c("CAR", "LPC", "PC", "SM")){
    cmps$adduct[i] <- "[M+H]+ ; [2M+H]+ ; [M+CHO2]- ; [2M+CHO2]-"
  } else if(cmps$class[i] %in% c("Cer", "Cer;O3", "Cer;O4", "HexCer", "LactCer")){
    cmps$adduct[i] <- paste(cmps$adduct[i], "; [M+CHO2]- ; [2M+CHO2]-")
  } else if(cmps$class[i] %in% c("DGMG")){
    cmps$adduct[i] <- "[M+NH4]+ ; [2M+NH4]+ ; [M-H]- ; [2M-H]- ; [M+CHO2]- ; [2M+CHO2]-"
  }
}
rm(cmps_RT, cmps_RT_IS, i, idx)
```


It is also required to define *metadata* for the database with the 
`make_metadata` function.

```{r}
metad <- make_metadata(source = "lipids", url = "",
                       source_version = "2", 
                       source_date = Sys.Date(),
                       organism = NA_character_)
```

With this information we can already create the `CompDb` database.

```{r}
db_file <- createCompDb(cmps, metadata = metad, 
                        path = "data/", dbFile = "CompDb_lipids_2.sqlite")
```

Below we load this database using the `CompDb` function in read-write mode 
because we want to add more information to this database later.

```{r}
(cdb <- CompDb(db_file, flags = RSQLite::SQLITE_RW))
rm(db_file, metad)
```

Next step is to create and `IonDb` database, which extends the `CompDb` by 
adding additional database tables to support ion annotations.

```{r}
idb <- IonDb(cdb)
```

Below we define a data frame with ion annotations and add that to the database 
with the `insertIon` function. The ion annotations are associated with the 
class of lipid. The column "compound_id" needs to contain the identifiers of 
the compounds to which the ion should be related to.

```{r iondb}
adds <- rbind(
  adducts("positive"),
  adducts("negative"),
  "[M+C2H8N]+" = c(
    "[M+C2H8N]+", 1, calculateMass("C2H8N"), "C2H8N", "H", 1, TRUE
  ),
  "[M-H-CO2]-" = c(
    "[M-H-CO2]-", 1, calculateMass("CO2")*(-1)-1.007276, 
    "CO2", "H", -1, FALSE
  ),
  "[M-H-H2O]-" = c(
    "[M-H-H2O]-", 1, calculateMass("H2O")*(-1)-1.007276, 
    "H2O", "H", -1, FALSE
  )
)
adds$mass_multi <- as.numeric(adds$mass_multi)
adds$mass_add <- as.numeric(adds$mass_add)
adds$positive <- as.logical(adds$positive)

for(i in seq(nrow(cmps))){
  x <- unlist(strsplit(cmps$adduct[i], " ; "))
  ion <- data.frame(
    compound_id = rep(cmps$compound_id[i], length(x)),
    ion_adduct = x,
    ion_mz = rep(NA, length(x)),
    ion_rt = rep(cmps$rtime[i], length(x))
  )
  for(j in seq(nrow(ion))){
    ion$ion_mz[j] <- mass2mz(cmps$exactmass[i], adds[x[j],])
  }
  idb <- insertIon(idb, ion)
}
rm(i, x, ion, j)
```


# Save the output

```{r save}
ionsdb <- ions(idb, c(
  "compound_id", "name", "ion_adduct", "ion_mz", "ion_rt", "class", 
  "t_min", "t_max", "m_min", "m_max"))
save(ionsdb, adds, file = "data/ionsdb.RData")
```


# Session info

```{r session-info}
Sys.time()-startpoint
devtools::session_info()
```

