---
title: "CompoundDb"
author: "Mar Garcia-Aloy"
output: 
  BiocStyle::html_document:
    toc: false
    number_sections: false
    toc_float: false
---

https://rformassspectrometry.github.io/CompoundDb/articles/create-compounddb.html#compdb-from-custom-data

```{r, include=FALSE}
startpoint <- Sys.time()
```

# Libraries

```{r, message=FALSE}
library(Rdisop)
library(CompoundDb)
library(MetaboCoreUtils)
```


# Main code

Create a `CompDb` database from manually defined annotations.  
Create a `data.frame` with compound annotations of analytical standards.   
The data needs to have the following information "compound_id", "name", "inchi", "inchikey", "formula", "exactmass", "synonyms".  
All columns except "compound_id" can also contain missing values.    
It is also possible to define additional columns.  

```{r}
cmps <- read.csv("data/compounds.csv")
cmps <- data.frame(
  compound_id = paste0(
    "C", formatC(seq(nrow(cmps)), width = nchar(nrow(cmps)), flag = "0")),
  name = cmps$compound,
  inchi = NA_character_,
  inchikey = NA_character_,
  formula = cmps$formula,
  exactmass = NA,
  synonyms = NA_character_,
  class = cmps$class,
  adduct = c("[M+H]+ ; [2M+H]+ ; [M-H]- ; [2M-H]-"), # LPE - LPS - PE - PS 
  # ¿Chol?
  rtime = cmps$rtime,
  std = cmps$standard,
  instd = cmps$IS,
  C = cmps$C,
  db = cmps$db
)
for(i in seq(nrow(cmps))){
  cmps$exactmass[i] <- getMolecule(cmps$formula[i])$exactmass
  if(cmps$class[i] %in% c("FFA", "LPA", "LPG", "LPI", "PA", "PG", "PI", "MG", "DG", "TG")){
    cmps$adduct[i] <- "[M+NH4]+ ; [2M+NH4]+ ; [M-H]- ; [2M-H]-"
  } else if(cmps$class[i] %in% c("MGDG", "DGDG")){
    cmps$adduct[i] <- "[M+NH4]+ ; [2M+NH4]+ ; [M+CHO2]- ; [2M+CHO2]-"
  } else if(cmps$class[i] %in% c("CAR", "LPC", "PC", "SM")){
    cmps$adduct[i] <- "[M+H]+ ; [2M+H]+ ; [M+CHO2]- ; [2M+CHO2]-"
  } else if(cmps$class[i] %in% c("Cer", "HexCer", "LactCer")){
    cmps$adduct[i] <- paste(cmps$adduct[i], "; [M+CHO2]- ; [2M+CHO2]-")
  }
}
rm(i)
```

It is also required to define *metadata* for the database with the `make_metadata` function.

```{r}
metad <- make_metadata(source = "lipids", url = "",
                       source_version = "1", 
                       source_date = Sys.Date(),
                       organism = NA_character_)
```

With this information we can already create the `CompDb` database.

```{r}
db_file <- createCompDb(cmps, metadata = metad, 
                        path = "data/", dbFile = "CompDb_lipids_1.sqlite")
```

Below we load this database using the `CompDb` function in read-write mode because we want to add more information to this database later.

```{r}
(cdb <- CompDb(db_file, flags = RSQLite::SQLITE_RW))
rm(db_file, metad)
```

Next step is to create and `IonDb` database, which extends the `CompDb` by adding additional database tables to support ion annotations.

```{r}
idb <- IonDb(cdb)
```

Below we define a data frame with ion annotations and add that to the database with the `insertIon` function. The ion annotations are associated with the class of lipid. The column "compound_id" needs to contain the identifiers of the compounds to which the ion should be related to.

```{r}
for(i in seq(nrow(cmps))){
  x <- unlist(strsplit(cmps$adduct[i], " ; "))
  ion <- data.frame(
    compound_id = rep(cmps$compound_id[i], length(x)),
    ion_adduct = x,
    ion_mz = rep(NA, length(x)),
    ion_rt = rep(cmps$rtime[i], length(x))
  )
  for(j in seq(nrow(ion))){
    ion$ion_mz[j] <- mass2mz(cmps$exactmass[i], x[j])
  }
  idb <- insertIon(idb, ion)
}
rm(i, x, ion, j)
```


# Save the output

```{r}
ionsdb <- ions(idb, c("compound_id", "name", "ion_adduct", 
                      "ion_mz", "ion_rt", "class"))
save(ionsdb, file = "data/ionsdb.RData")
```


# Session info

```{r}
Sys.time()-startpoint
devtools::session_info()
```

