---
title: "Processing"
author: "Mar Garcia-Aloy"
output: 
  BiocStyle::html_document:
    toc: true
    number_sections: false
    toc_float: true
---

```{r startpoint, include = FALSE}
startpoint <- Sys.time()
```

In this document I'm going to import DDA data on study samples in order to select those peaks that have at least 1 MS2 spectra.  
Fragmentation spectra has obtained using data dependent acquisition (DDA) method. In this method the top N most intense m/z values from a MS1 scan are selected for fragmentation in the next N scans before the cycle starts again.  
The acquired data contains both MS1 and MS2 spectra.

https://bioconductor.org/packages/release/bioc/vignettes/xcms/inst/doc/xcms-lcms-ms.html#LC-MSMS_data_analysis_with_xcms

# Libraries

```{r libraries, message = FALSE, warning = FALSE}
library(xcms)
```


# Data import

```{r import, eval = FALSE}
for(s in c("tissues", "maturation")){
  for(p in c("POS", "NEG")){
    fln <- list.files(paste0("data/", s, "/", p, "_DDA_mzML/"))
    fln <- fln[!grepl("MIX|QC|entire|STD", fln)]
    mtd <- gsub("Rep_", "rep", fln)
    mtd <- gsub("Pt_", "pt", mtd)
    mtd <- gsub("lipidgrape_tissues", "lipidgrape", mtd)
    if(s == "tissues"){
      mtd <- gsub("pt11_", "", mtd)
    }
    mtd <- strsplit(gsub(".mzML", "", mtd), "_")
    mtd <- data.frame(do.call(rbind, mtd))
    colnames(mtd) <- c("order", "project", "class", "rep", "p", "mode")
    injections <- data.frame(cbind(fln, mtd))
    
    # Read the data
    dda_data <- readMSData(
      paste0("data/", s, "/", p, "_DDA_mzML/" , injections$fln),
      pdata = as(AnnotatedDataFrame(injections),
                 "NAnnotatedDataFrame"), 
      mode = "onDisk")
    
    assign(paste("dda_data", s, p, sep = "_"), dda_data)
    save(dda_data, file = paste0("data/tmp/dda_data_", s, "_", p, ".RData"))
  }
}
rm(injections, fln, mtd, dda_data) #s, p
```

```{r load}
for(s in c("tissues", "maturation")){
  for(p in c("POS", "NEG")){
    load(paste0("data/tmp/dda_data_", s, "_", p, ".RData"))
    assign(paste("dda_data", s, p, sep = "_"), dda_data)
  }
}
```


# Peak detection & grouping

Here we perform the chromatographic peak detection on the MS level 1 data, followed by peak grouping.

```{r peak-detection}
for(s in c("tissues", "maturation")){
  for(p in c("POS", "NEG")){
    dda_xdata <- get(paste("dda_data", s, p, sep = "_"))
    if(p == "NEG"){
      int <- 1e4
    } else if(p == "POS"){
      int <- 1e5
    }
    dda_xdata <- findChromPeaks(
      dda_xdata, 
      param = CentWaveParam(ppm = 20,
                            peakwidth = c(5, 40),
                            prefilter = c(5, int),
                            snthresh = 5,
                            noise = 100,
                            mzdiff = 0.001,
                            integrate = 2)
    )
    register(SerialParam())
    dda_xdata <- refineChromPeaks(
      dda_xdata, 
      param = FilterIntensityParam(nValues = 5, threshold = int)
    )
    dda_xdata <- groupChromPeaks(
      dda_xdata, 
      param = PeakDensityParam(sampleGroups = dda_xdata$class,
                               minFraction = 1,
                               binSize = 0.02,
                               bw = 3)
    )
    chromPeaks(dda_xdata)
    featureDefinitions(dda_xdata)
    assign(paste("dda_xdata", s, p, sep = "_"), dda_xdata)
  }
}
rm(int, dda_xdata)
```

# MS1 and MS2 data integration

The advantage of LC-MS/MS data is that (MS1) ions are fragmented and the corresponding MS2 spectra measured. Thus, for some of the ions (identified as MS1 chromatographic peaks) MS2 spectra are available.   
Below we firstly extract the MS2 spectra to the object `dda_spectra` and the features detected in the MS1 data to the object `ft`.

```{r ms2-integration1}
for(s in c("tissues", "maturation")){
  for(p in c("POS", "NEG")){
    dda_xdata <- get(paste("dda_xdata", s, p, sep = "_"))
    ft <- as.data.frame(featureDefinitions(dda_xdata))
    assign(paste("ft", s, p, sep = "_"), ft)
  }
}
rm(ft, dda_xdata)
```


```{r exclude-ms2}
# Exclude noisy MS2:
ms2_exclude <- read.csv("data/MS2_exclude.csv")

for(s in c("tissues", "maturation")){
  for(p in c("POS", "NEG")){
    dda_xdata <- get(paste("dda_xdata", s, p, sep = "_"))
    dda_spectra <- chromPeakSpectra(
      dda_xdata, msLevel = 2L, return.type = "Spectra")
    tmp_ms2_exclude <- ms2_exclude[grep(p, ms2_exclude$File),]
    if(s == "tissues"){
      tmp_ms2_exclude <- tmp_ms2_exclude[grepl("tissues", tmp_ms2_exclude$File),]
    } else if(s == "maturation"){
      tmp_ms2_exclude <- tmp_ms2_exclude[!grepl("tissues", tmp_ms2_exclude$File),]
    }
    
    if(nrow(tmp_ms2_exclude) > 0){
      idx <- rep(NA, nrow(tmp_ms2_exclude))
      for(i in seq(length(idx))){
        x <- which(
          basename(
            dda_spectra@backend@spectraData@listData$dataOrigin
          ) == tmp_ms2_exclude$File[i] & 
            dda_spectra@backend@spectraData@listData$scanIndex == tmp_ms2_exclude$scan[i]
        )
        if(length(x) > 0){
          idx[i] <- x
        }
      }
      idx <- idx[!is.na(idx)]
      dda_spectra <- dda_spectra[-idx]
    }
    
    assign(paste("dda_spectra", s, p, sep = "_"), dda_spectra)
  }
}
rm(dda_xdata, dda_spectra, idx, i, x, ft, ms2_exclude, tmp_ms2_exclude)
```

Finally, for each feature we extract all MS2 spectra that were associated with the corresponding feature using the ID of the corresponding chromatographic peaks. We annotate how many MS2 spectra do we have for each feature in order to keep only those features with at least 1 MS2 spectra.

```{r ms2-integration2}
for(s in c("tissues", "maturation")){
  for(p in c("POS", "NEG")){
    ft <- get(paste("ft", s, p, sep = "_"))
    dda_spectra <- get(paste("dda_spectra", s, p, sep = "_"))
    dda_xdata <- get(paste("dda_xdata", s, p, sep = "_"))
    ft$n_MS2 <- NA
    for(i in seq(nrow((ft)))){
      i_id <- rownames(chromPeaks(dda_xdata)[ft[i, "peakidx"][[1]],])
      i_spectra <- dda_spectra[dda_spectra$peak_id %in% i_id]
      ft$n_MS2[i] <- length(i_spectra)
    }
    assign(paste("ft", s, p, sep = "_"), ft)
  }
}

rm(s, p, ft, dda_spectra, dda_xdata, i, i_id, i_spectra)
```


# Save data

```{r save}
save.image("data/RData.RData")
```


# Session information

```{r session}
Sys.time()-startpoint
devtools::session_info()
```

