---
title: "Identification"
author: "Mar Garcia-Aloy"
output: 
  BiocStyle::html_document:
    toc: true
    number_sections: false
    toc_float: true
---

# Libraries

```{r libraries, message = FALSE, warning = FALSE}
library(xcms)
library(MsFeatures)
library(MetaboCoreUtils)
library(MetaboAnnotation)
library(magrittr)
library(OrgMassSpecR)
library(CompoundDb)
library(Rdisop)
library(MsCoreUtils)
library(Spectra)

.ppm <- function(x, ppm = 10) {
  ppm * x / 1e6
}
matchWithPpm <- function(x, y, ppm = 0) {
  lapply(x, function(z, ppm) {
    which(abs(z - y) <= (.ppm(z, ppm)) + 1e-9)
  }, ppm = force(ppm))
}
max_peak <- function(x, ...) {
  x[which.max(x[, 2]), , drop = FALSE]
}
```

# Data import

```{r import}
load("data/RData.RData")
load("data/ionsdb.RData")
load("data/colors.RData")
load("data/MS2_STDmix.RData")
db_file <- "data/CompDb_lipids_1.sqlite"
startpoint <- Sys.time()
```

# Isotopologues

We firstly want to detect isotopes using the function `isotopologues`.   
This function identifies isotopes in a given MS spectrum. However, our data is LC-MS. Therefore, first of all we group features according to retention time (http://sneumann.github.io/xcms/articles/LC-MS-feature-grouping.html), in order to have something similar to a pseudo-MS spectrum, and then within each feature group we apply the function.  


```{r isotopes}
for(s in c("tissues", "maturation")){
  for(p in c("POS", "NEG")){
    dda_xdata <- get(paste("dda_xdata", s, p, sep = "_"))
    ft <- get(paste("ft", s, p, sep = "_"))
    
    # feature grouping:
    dda_xdata <- groupFeatures(dda_xdata, param = SimilarRtimeParam(10))
    ft <- cbind(ft, featureGroups(dda_xdata))
    colnames(ft)[ncol(ft)] <- "FG"
    
    # calculate mean-intensities of each feature:
    dt <- data.frame(t(featureValues(dda_xdata, value = "into")))
    all(colnames(dt) == rownames(ft))
    ft$into <- apply(dt, 2, mean, na.rm = TRUE)
    
    # identify isotopes:
    ft$isotope <- NA
    ft$is_isotope <- FALSE
    ft <- ft[order(ft$mzmed),]
    xi <- 0
    for(j in unique(ft$FG)){ # within FG "j"....
      tmp <- ft[ft$FG == j,]
      if(nrow(tmp) > 1){ # if there is >1feature in FG "j", apply "isotopologues"
        isos <- isotopologues(as.data.frame(tmp[,c("mzmed", "into")]), 
                              ppm = 10)
        if(length(isos) > 0){ # if there are isotopes detected, annotate them
          for(i in seq_along(isos)){
            z <- isos[[i]]
            xi <- xi + 1
            z <- which(rownames(ft) %in% rownames(tmp)[z])
            ft$isotope[z] <- paste0(
              "i", formatC(xi, width = 2, flag = "0")
            )
            ft$is_isotope[z[-1]] <- TRUE
          }
        }
      }
    }
    ft$dataset <- paste(s, p, sep = "_")
    assign(paste("dda_xdata", s, p, sep = "_"), dda_xdata)
    assign(paste("ft", s, p, sep = "_"), ft)
  }
}
rm(dda_xdata, ft, dt, xi, j, tmp, isos, i, z) # fts_idx
```


# Annotation

```{r annotation}
ionsdb$ion_rt <- ionsdb$ion_rt*60
for(s in c("tissues", "maturation")){
  for(p in c("POS", "NEG")){
    ft <- get(paste("ft", s, p, sep = "_"))
    ft <- ft[ft$n_MS2 > 0,] # select features with at least 1  MS2
    ft <- ft[!ft$is_isotope,] # select non-isotopic features
    # get the ions from polarity "p":
    if(p == "POS"){
      tmp_ionsdb <- ionsdb[ionsdb$ion_adduct %in% adducts("positive")[,"name"],]
    } else if(p == "NEG"){
      tmp_ionsdb <- ionsdb[ionsdb$ion_adduct %in% adducts("negative")[,"name"],]
    }
    # start matching based on mz & RT:
    param <- MzRtParam(ppm = 5, toleranceRt = 5)
    pks_match <- matchMz(ft, tmp_ionsdb, param, 
                         mzColname = c("mzmed", "ion_mz"),
                         rtColname = c("rtmed", "ion_rt"))
    pks <- as.data.frame(matchedData(pks_match))
    pks_rt <- pks[!is.na(pks$target_compound_id),]
    # unify multiple matches:
    dpl <- unique(gsub("\\..*", "", rownames(pks_rt)[grep(
      "\\.", rownames(pks_rt))]))
    for(i in dpl){
      idx <- c(which(rownames(pks_rt) == i), grep(
        paste0(i, "."), rownames(pks_rt)))
      tmp <- pks_rt[idx,]
      pks_rt$target_compound_id[idx[1]] <- paste(
        pks_rt$target_compound_id[idx], collapse = "-")
      pks_rt$target_ion_adduct[idx[1]] <- paste(
        unique(pks_rt$target_ion_adduct[idx]), collapse = "-")
      pks_rt$target_name[idx[1]] <- paste(
        unique(pks_rt$target_name[idx]), collapse = "-")
      pks_rt <- pks_rt[-idx[-1],]
    }
    
    # for those un-annotated features, match based only on mz:
    if(nrow(pks_rt) < nrow(ft)){
      param <- MzParam(ppm = 5)
      pks_match <- matchMz(ft[!rownames(ft) %in% rownames(pks_rt),], 
                           tmp_ionsdb, param, 
                           mzColname = c("mzmed", "ion_mz"))
      pks <- as.data.frame(matchedData(pks_match))
      # unify multiple matches:
      dpl <- unique(gsub("\\..*", "", rownames(pks)[
        grep("\\.",rownames(pks))]))
      for(i in dpl){
        idx <- c(which(rownames(pks) == i), grep(
          paste0(i, "."), rownames(pks)))
        tmp <- pks[idx,]
        pks$target_compound_id[idx[1]] <- paste(
          pks$target_compound_id[idx], collapse = "-")
        pks$target_ion_adduct[idx[1]] <- paste(
          unique(pks$target_ion_adduct[idx]), collapse = "-")
        pks$target_name[idx[1]] <- paste(unique(
          pks$target_name[idx]), collapse = "-")
        pks <- pks[-idx[-1],]
      }
      pks$score_rt <- NA
      colnames(pks)[ncol(pks)-1] <- "ppm_error"
      pks <- rbind(pks_rt, pks)
    }
    
    pks <- pks[order(rownames(pks)),]
    pks <- pks[,c("mzmed", "rtmed", "n_MS2", "FG", "into", "isotope", 
                  "is_isotope", "target_compound_id", "target_ion_adduct", 
                  "target_name", "target_class", "ppm_error", "score_rt")]
    pks$FT <- rownames(pks)
    pks$rtmed <- pks$rtmed/60
    if(s == "tissues" & p == "POS"){
      head(pks)
    }
    print(paste(s, p))
    pks$dataset <- paste(s, p, sep = "_")
    assign(paste("pks", s, p, sep = "_"), pks)
    pks <- pks[!is.na(pks$target_compound_id),]
    pks <- pks[!grepl("-", pks$target_compound_id),]
    pks <- pks[order(pks$target_compound_id, pks$rtmed),]
    pander::pandoc.table(
      pks[,c("mzmed", "rtmed", "target_compound_id", "target_name", "n_MS2", "score_rt")],
      style = "rmarkdown", split.tables = Inf, row.names = FALSE,
      digits = c(7, 4, rep(0, 4)))
  }
}
rm(ft, tmp_ionsdb, param, pks_match, pks_rt, dpl, i, idx, tmp, pks)
```


# Workflow

```{r workflow}
for(s in c("tissues", "maturation")){
  for(p in c("POS", "NEG")){
    dda_xdata <- get(paste("dda_xdata", s, p, sep = "_"))
    ft <- get(paste("ft", s, p, sep = "_"))
    pks <- get(paste("pks", s, p, sep = "_"))
    print(paste(s, p))
    print(paste("Detected features in MS1 data:", 
                nrow(featureDefinitions(dda_xdata))))
    print(paste("Non-isotopic MS1 features:", sum(!ft$is_isotope)))
    pks <- pks[!pks$is_isotope,]
    print(paste("Non-isotopic MS1 features with at least one MS2 spectra:", nrow(pks)))
    pks <- pks[!is.na(pks$target_compound_id),]
    pks <- pks[!grepl("-", pks$target_compound_id),]
    print(paste("Annotated features:", nrow(pks)))
    print(paste("Unique compounds:", length(unique(pks$target_compound_id))))
    print("################################") 
  }
}
rm(dda_xdata, ft, pks)
```


# Table

```{r table}
pks <- rbind(pks_tissues_POS, pks_maturation_POS,
             pks_tissues_NEG, pks_maturation_NEG)
###########################################################################
pks <- pks[!is.na(pks$score_rt) & !grepl("-", pks$target_compound_id),]
#pks <- pks[!is.na(pks$ppm_error) & !grepl("-", pks$target_compound_id),]
###########################################################################
cmps <- compounds(CompDb(db_file), c(
  "compound_id", "name", "formula", "exactmass", "class", "std", "instd", "C", "db"))
cmps <- cmps[cmps$compound_id %in% pks$target_compound_id,]
cmps$RT <- NA
cmps$tissues <- ""
cmps$maturation <- ""
for(i in seq(nrow(cmps))){
  tmp <- pks[pks$target_compound_id == cmps$compound_id[i],]
  cmps$RT[i] <- mean(tmp$rtmed)
  cmps$tissues[i] <- paste(tmp$target_ion_adduct[
    grep("tissues", tmp$dataset)], collapse = "; ")
  cmps$maturation[i] <- paste(tmp$target_ion_adduct[
    grep("maturation", tmp$dataset)], collapse = "; ")
}
cmps$class <- factor(cmps$class, levels = c(
  "FFA", "CAR", "LPA", "LPC", "LPE", "LPG", "LPI", "LPS", 
  "PA", "PC", "PE", "PG", "PI","PS",
  "SM", "Cer", "HexCer", "LactCer", 
  "MGDG", "DGDG", "SQDG", "MG", "DG", "TG"))
cmps <- cmps[order(-cmps$instd, cmps$class, cmps$C, cmps$db, cmps$RT),]
cmps$cmp <- paste0("C", formatC(seq(nrow(cmps)), width = nchar(nrow(cmps)), 
                                flag = "0"))
nrow(cmps)
for(s in c("tissues", "maturation")){
  print(s)
  print(paste("Total compounds:", sum(cmps[,s] != "")))
  print(paste("Identified compounds:", sum(cmps[cmps$class != "unk", s] != "")))
  print(paste("Unknown compounds:", sum(cmps[cmps$class == "unk", s] != "")))
  print("###########################################")
}
cmps$RT <- sprintf("%.2f", cmps$RT)
DT::datatable(
  cmps[,c("cmp", "name", "formula", "RT", "tissues", "maturation")],
  rownames = FALSE,
  options = list(pageLength = nrow(cmps)#, dom = "t"
  ))
rm(i, tmp)
write.csv(cmps[,c("cmp", "name", "formula", "RT", "tissues", "maturation")], 
          "output/identifications.csv")
```


# Figures

## Venn Diagram

```{r venn}
tissues <- unique(c(pks_tissues_NEG$target_name, pks_tissues_POS$target_name))
maturat <- unique(c(pks_maturation_NEG$target_name, pks_maturation_POS$target_name))
tissues <- tissues[!is.na(tissues)]
maturat <- maturat[!is.na(maturat)]

VennDiagram::draw.pairwise.venn(
  area1 = length(tissues),
  area2 = length(maturat),
  cross.area = length(tissues[tissues %in% maturat]),
  category = c("Tissues", "Ripening")
)
```


## Classes of compounds

```{r classes, eval = TRUE}
cmps$class_super <- NA
cmps$class_super[which(
  cmps$class %in% c("LPC", "LPE", "LPG", "LPI", "LPS",
                    "PA", "PC", "PE", "PI", "PG", "PS"))] <- "(lyso)PL"
cmps$class_super[which(cmps$class %in% 
                         c("MGDG", "DGDG", "SQDG", "DG", "TG"))] <- "GL"
cmps$class_super[which(cmps$class %in% c("SM", "Cer", "HexCer", "LactCer"))] <- "SP"
cmps$class_super[cmps$class %in% c("CAR", "FFA")] <- "others"
unique(cmps$class[is.na(cmps$class_super)])
cmps$class <- as.character(cmps$class)
webr::PieDonut(cmps, ggplot2::aes(pies = class_super, donuts = class), 
               showPieName = FALSE)
```


## Fatty acyl chains

```{r fa-chains, eval = TRUE}
idx <- which(cmps$instd == 0) # discard IS
cl <- unique(cmps$class[idx])
fa <- gsub(".*\\ ", "", cmps$name[idx]) # remove class from name
fa <- unique(unlist(strsplit(fa, "_"))) # separate FA chains & get unique values
links <- data.frame(expand.grid(cl, fa))
colnames(links) <- c("source", "target")
links$value <- 0
for(i in seq(nrow(links))){
  links$value[i] <- length(grep(
    links$target[i], unlist(
      strsplit(cmps$name[cmps$class == links$source[i]], "_"))))
}
links <- links[links$value > 0,]
nodes <- data.frame(
  name = c(as.character(links$source), 
           as.character(links$target)) %>% unique()
)
links$IDsource <- match(links$source, nodes$name)-1 
links$IDtarget <- match(links$target, nodes$name)-1
networkD3::sankeyNetwork(Links = links, Nodes = nodes,
                         Source = "IDsource", Target = "IDtarget",
                         Value = "value", NodeID = "name", 
                         sinksRight = FALSE)
rm(cl, fa, links, i, nodes)
```



## KMD

```{r kmd, eval = TRUE}
ref <- c(12, 1, 16, 14, 31, 32)
names(ref) <- c("C", "H", "O", "N", "P", "S")
cmps$NM <- NA
for(i in seq(nrow(cmps))){
  if(cmps$formula[i] != "X"){
    myform <- countElements(cmps$formula[i])
    pippo <- ref[names(myform[[1]])]
    cmps$NM[i] <- t(pippo) %*% myform[[1]]
  } else {
    cmps$NM[i] <- floor(cmps$exactmass[i])
  }
}
cmps$KMD_H <- cmps$NM - (
  cmps$exactmass * ref["H"] / 
    MonoisotopicMass(formula = ListFormula("H")))
myform <- countElements("CH2")
pippo <- ref[names(myform[[1]])]
cmps$KMD_CH2 <- cmps$NM - (
  cmps$exactmass * as.numeric(t(pippo) %*% myform[[1]]) / 
    MonoisotopicMass(formula = ListFormula("CH2")))


idx <- which(cmps$instd == 0) # discard IS
plot(cmps$KMD_CH2[idx], cmps$KMD_H[idx], xlab = "KMD(CH2)", ylab = "KMD(H)", 
     col = "white")
# indicate the C:db (for the unknowns indicate the name):
text(cmps$KMD_CH2, cmps$KMD_H, paste(cmps$C, cmps$db, sep = ":"), 
     col = col_cmps[cmps$class], cex = 0.8)

# define colors by number of double bonds:
mycols <- inlmisc::GetColors(max(cmps$db) + 1, "discrete rainbow")
names(mycols) <- c(seq(max(cmps$db)), 0)

# get the data from standards NOT included in study compounds:
stds <- compounds(CompDb(db_file), c(
  "compound_id", "name", "formula", "exactmass", "rtime", "class", "std", "instd", "C", "db"))
stds <- stds[stds$std == 1 & stds$instd == 0, ]
stds <- stds[!stds$name %in% cmps$name, ]
stds <- stds[stds$rtime > 0, ]
stds$exactmass <- NA
stds$NM <- NA
for(i in seq(nrow(stds))){
  stds$exactmass[i] <- Rdisop::getMolecule(stds$formula[i])$exactmass
  myform <- countElements(stds$formula[i])
  pippo <- ref[names(myform[[1]])]
  stds$NM[i] <- t(pippo) %*% myform[[1]]
}
stds$KMD_H <- stds$NM - (
  stds$exactmass * ref["H"] / 
    MonoisotopicMass(formula = ListFormula("H")))
stds$RT <- stds$rtime

cmps$class <- factor(cmps$class, levels = c(
  "FFA", "CAR", "LPA", "LPC", "LPE", "LPG", "LPI", "LPS", 
  "PA", "PC", "PE", "PG", "PI","PS",
  "SM", "Cer", "HexCer", "LactCer", 
  "MGDG", "DGDG", "SQDG", "MG", "DG", "TG"))
cl <- unique(cmps$class[cmps$instd == 0])
pdf("output/KMD.pdf", paper = "a4", height = 4*3, width = 2*4)
par(mfrow = c(4, 2), mar = c(4, 3.5, 2, 0.5))
for(i in seq(length(cl))){
  # get the data from compounds of class "i":
  tmp <- rbind(
    cmps[cmps$class == cl[i], c("compound_id", "class", "name", "RT", 
                                "KMD_H", "std", "instd", "C", "db")],
    stds[stds$class == cl[i], c("compound_id", "class", "name", "RT", 
                                "KMD_H", "std", "instd", "C", "db")])
  tmp <- tmp[tmp$instd == 0, ]
  if(nrow(tmp) > 1){
    # define ylims:
    if(cl[i] %in% c("CAR", "LPC", "PE", "PG", 
                    "SM", "Cer", "HexCer", "LactCer", "MGDG", "DGDG", 
                    "MG", "DG")){
      y <- 0.06
    } else if(cl[i] %in% c("FFA", "PA")){
      y <- 0.07
    }else if(cl[i] %in% c("TG")){
      y <- 0.1
    } else {
      y <- 0.05
    }
    # do the plot:
    plot(tmp$RT, tmp$KMD_H, col = "white", xlab = "", ylab = "", bty="l",
         ylim = c(min(tmp$KMD_H) - y, max(tmp$KMD_H) + 0.05))
    points(tmp$RT, tmp$KMD_H, pch = 19, cex = 2, 
           col = mycols[as.character(tmp$db)])
    idx <- which(tmp$std == 1) # compounds with STDs
    points(tmp$RT[idx], tmp$KMD_H[idx], pch = 3, cex = 2)
    idx <- which(tmp$name %in% cmps$name) # study compounds
    points(tmp$RT[idx], tmp$KMD_H[idx], pch = 1, cex = 3)
    abline(h=unique(round(tmp$KMD_H,2)), lty =2 , col = "grey")
    tmp$Cdb <- paste(tmp$C, tmp$db, sep = ":")
    if(any(duplicated(tmp$Cdb))){
      d <- tmp$Cdb[duplicated(tmp$Cdb)]
      x <- tmp$RT[!duplicated(tmp$Cdb)]
      y <- tmp$KMD_H[!duplicated(tmp$Cdb)]
      z <- tmp$Cdb[!duplicated(tmp$Cdb)]
      for(j in seq(length(d))){
        x[which(z == d[j])] <- mean(as.numeric(tmp$RT[tmp$Cdb == d[j]]))
      }
      text(x, y, z, pos = 1, offset = 1)
    } else {
      text(tmp$RT, tmp$KMD_H, tmp$Cdb, pos = 1, offset = 1)
    }
    title(cl[i], line = 0)
    title(xlab = "RT (min)", ylab = "KMD(H)", line = 2.5)
    # add diagonal lines:
    for(j in unique(tmp$db[duplicated(tmp$db)])){
      idx <- which(tmp$db == j)
      # only if there is >1 non-isomeric compound with "j" double bonds:
      if(length(unique(round(tmp$KMD_H[idx], 2))) > 1){
        lines(as.numeric(tmp$RT[idx]), 
              predict(lm(tmp$KMD_H[idx]~as.numeric(tmp$RT[idx]))), lwd = 4, 
              col = paste0(mycols[as.character(j)], "60"))
      }
    } # close db "j"
  } # close "if(nrow(tmp) > 0)"
} # close cl "i"
dev.off()


# Equivalent carbon number
cmps$ECN <- cmps$C - 2*cmps$db
stds$ECN <- stds$C - 2*stds$db

cmps$RT <- as.numeric(cmps$RT)
cmps$rRT <- cmps$RT/max(cmps$RT)
cmps$rECN <- cmps$ECN/max(cmps$ECN)
mylm <- lm(rRT ~ poly(rECN, degree = 2), data = cmps)
newdata <- data.frame(
  rECN = seq(min(cmps$rECN) - 0.01, max(cmps$rECN) + 0.01, by = 0.01)
)
plot(cmps$rECN, cmps$rRT, bty="l", pch = 16, col = col_cmps[cmps$class],
     xlab = "Relative ECN", ylab = "Relative RT")
lines(x = newdata$rECN, 
      y = predict(lm(rRT ~ poly(rECN, degree = 2), data = cmps), 
                  newdata = newdata))
text(cmps$rECN, cmps$rRT, paste0(cmps$class, cmps$C, ":", cmps$db), 
     cex = 0.5, pos = 1)

pdf("output/ECN.pdf", paper = "a4", height = 4*3, width = 2*4)
par(mfrow = c(4, 2), mar = c(4, 3.5, 2, 0.5))
for(i in seq(length(cl))){
  # get the data from compounds of class "i":
  tmp <- rbind(
    cmps[cmps$class == cl[i], c("compound_id", "class", "name", "RT", 
                                "KMD_H", "std", "instd", "C", "db", "ECN")],
    stds[stds$class == cl[i], c("compound_id", "class", "name", "RT", 
                                "KMD_H", "std", "instd", "C", "db", "ECN")])
  tmp <- tmp[tmp$instd == 0, ]
  tmp$RT <- as.numeric(tmp$RT)
  tmp$ECN <- as.numeric(tmp$ECN)
  tmp <- tmp[order(tmp$ECN),]
  tmp$rRT <- tmp$RT/max(tmp$RT)
  tmp$rECN <- tmp$ECN/max(tmp$ECN)
  if(length(unique(tmp$ECN)) > 2){
    tmp_lm <- lm(rRT ~ poly(rECN, degree = 2), data = tmp)
    newdata <- data.frame(
      rECN = seq(min(tmp$rECN) - 0.01, max(tmp$rECN) + 0.01, by = 0.01)
    )
    plot(tmp$rECN, tmp$rRT, bty="l", pch = 16,
         xlab = "Relative ECN", ylab = "Relative RT")
    lines(x = newdata$rECN, 
          y = predict(lm(rRT ~ poly(rECN, degree = 2), data = tmp), 
                      newdata = newdata))
    text(tmp$rECN, tmp$rRT, paste0(tmp$C, ":", tmp$db), 
     cex = 0.5, pos = 1)
    title(cl[i], line = 0)
  } # close "if(length(unique(tmp$ECN)) > 2)"
} # close cl "i"
dev.off()

rm(ref, i, myform, pippo, mycols, cl, tmp, y, idx, j)
```


## MS2 spectra

```{r ms2, eval = FALSE}
cdb <- CompDb(db_file, flags = RSQLite::SQLITE_RW)
# join data from study compounds & 
# standards from the classes included in study compounds
stds <- stds[stds$class %in% unique(cmps$class[cmps$instd == 0]),]
cmps$study <- 1
stds$study <- 0
stds$tissues <- ""
stds$maturation <- ""
cmps <- rbind(
  cmps[,c("compound_id", "name", "class", "RT", "formula", 
          "tissues", "maturation", "std", "instd", "study")], 
  stds[,c("compound_id", "name", "class", "RT", "formula", 
          "tissues", "maturation", "std", "instd", "study")])
cmps <- cmps[cmps$instd == 0, ]
cmps$class <- factor(cmps$class, levels = c(
  "FFA", "CAR", "LPA", "LPC", "LPE", "LPG", "LPI", "LPS", 
  "PA", "PC", "PE", "PG", "PI","PS",
  "SM", "Cer", "HexCer", "LactCer", 
  "MGDG", "DGDG", "SQDG", "MG", "DG", "TG"))
cmps$RT <- as.numeric(cmps$RT)
cmps <- cmps[order(cmps$class, cmps$name, cmps$RT, cmps$compound),]


# function to annotate the MS2 fragments:
ms2_ann <- function(mz, i, add, nl){
  sps <- data.frame(cbind(mz = mz, i = i))
  sps$i <- sps$i/max(sps$i)
  # annotate fragments:
  sps$ann <- NA
  for(j in seq(nrow(nl))){
    idx <- which(MsCoreUtils::between(
      sps$mz,
      as.numeric(mass2mz(mz2mass(i_mz, add), nl[j,])) + 0.01 * c(-1,1)
    ))
    if(length(idx) > 0){
      sps$ann[idx] <- paste(sps$ann[idx], rownames(nl)[j])
    }
  }
  sps$ann <- gsub("NA ", "", sps$ann)
  return(sps)
}

# adducts of neutral losses:
nl <- data.frame(rbind(
  c("[M+H]+", 1.007276, TRUE),
  c("[M+H-H2O]+", (getMolecule("H2O")$exactmass*(-1))+1.007276, TRUE),
  c("[M+H-C2H2]+", (getMolecule("C2H2")$exactmass*(-1))+1.007276, TRUE),
  c("[M+H-hexose]+", (getMolecule("C6H10O5")$exactmass*(-1))+1.007276, TRUE),
  c("[M+H-hexose-H2O]+", (getMolecule("C6H12O6")$exactmass*(-1))+1.007276, TRUE),
  c("[M+H-PA]+", (getMolecule("H3PO4")$exactmass*(-1))+1.007276, TRUE),
  c("[M+H-mPA]+", (getMolecule(addElements("H3PO4", "CH2"))$exactmass*(-1))+1.007276, TRUE),
  c("[M+H-dmPA]+", (getMolecule(addElements("H3PO4", "CH2CH2"))$exactmass*(-1))+1.007276, TRUE),
  c("[M+H-PA-C2H2]+", (getMolecule(addElements("H3PO4", "C2H2"))$exactmass*(-1))+1.007276, TRUE),
  c("[M+H-157.0504]+", 157.0504*(-1)+1.007276, TRUE),
  c("[M+H-PE]+", (getMolecule("C2H8NO4P")$exactmass*(-1))+1.007276, TRUE),
  c("[M+NH4-C3H9O6]+", (getMolecule("C3H9O6")$exactmass*(-1))+getMolecule("NH4")$exactmass, TRUE),
  c("[M+H-PI]+", (getMolecule("C6H13O9P")$exactmass*(-1))+1.007276, TRUE),
  c("[M-H]-", -1.007276, FALSE),
  c("[M-CH3]-", (getMolecule("CH3")$exactmass*(-1)), FALSE),
  c("[M-H-hexose]-", (getMolecule("C6H10O5")$exactmass*(-1))-1.007276, TRUE),
  c("[M-H-hexose-H2O]-", (getMolecule("C6H12O6")$exactmass*(-1))-1.007276, TRUE)
))
# add losses of fatty-acyl chains:
C <- seq(from = 14, to = 24, by = 1)
db <- seq(from = 0, to = 3, by = 1)
sn <- paste(expand.grid(C, db)[,"Var1"], expand.grid(C, db)[,"Var2"], 
            sep = ":")
sn <- cbind(sn, 
            paste0("C", gsub(":.*", "", sn), 
                   "H", as.numeric(gsub(":.*", "", sn))*2 - 
                     2*as.numeric(gsub(".*:", "", sn)), 
                   "O2"))
colnames(sn) <- c("sn", "formula")
for(i in seq(nrow(sn))){
  nl <- rbind(
    nl,
    # [M+H-18:2-H2O]+
    c(paste0("[M+H-", sn[i, "sn"],"-H2O]+"), getMolecule(sn[i, "formula"])$exactmass*(-1)+1.007276, TRUE),
    # [M-H-18:2-H2O]-
    c(paste0("[M-H-", sn[i, "sn"],"-H2O]-"), getMolecule(sn[i, "formula"])$exactmass*(-1)-1.007276, FALSE),
    # [M-H-18:2-H2O-glycerol]-
    c(paste0("[M-H-", sn[i, "sn"],"-H2O-glycerol]-"), getMolecule(subtractElements(addElements(sn[i, "formula"], "C3H8O3"), "H2O"))$exactmas*(-1)-1.007276, FALSE),
    # [M-H-18:2-H2O-inositol]-
    c(paste0("[M-H-", sn[i, "sn"],"-H2O-inositol]-"), getMolecule(subtractElements(addElements(sn[i, "formula"], "C6H12O6"), "H2O"))$exactmas*(-1)-1.007276, FALSE),
    # [M+H-18:2]+
    c(paste0("[M+H-", sn[i, "sn"],"]+"), getMolecule(subtractElements(sn[i, "formula"], "H2O"))$exactmass*(-1)+1.007276, TRUE),
    # [M-H-18:2]-
    c(paste0("[M-H-", sn[i, "sn"],"]-"), getMolecule(subtractElements(sn[i, "formula"], "H2O"))$exactmass*(-1)-1.007276, FALSE)
  )}
colnames(nl) <- c("name", "mass_add", "positive")
nl$mass_add <- as.numeric(nl$mass_add)
nl$mass_multi <- 1
nl$charge <- 1
nl$formula_add <- "X"
nl$formula_sub <- "X"
nl <- nl[, c("name", "mass_multi", "mass_add", "formula_add", "formula_sub", 
             "charge", "positive")]
rownames(nl) <- nl$name

# diagnostic ions:
ions <- data.frame(rbind(
  c(mass2mz(getMolecule("C5H14NO4P")$exactmass, "[M+H]+"), "[PC+H]+", TRUE)
))
colnames(ions) <- c("mz", "name", "positive")
for(i in seq(nrow(sn))){
  ions <- rbind(
    ions,
    c(mass2mz(getMolecule(sn[i, "formula"])$exactmass, "[M-H]-"), 
      paste0("[",sn[i, "sn"], "-H]-"), FALSE))
}
ions$mz <- as.numeric(ions$mz)

pdf("output/MS2.pdf", paper = "a4", height = 4*3, width = 2*4)
par(mfrow = c(3, 2), mar = c(4, 4, 3, 0.5))
for(i in seq(nrow(cmps))){
  for(p in c("POS", "NEG")){
    if(p == "POS"){
      tmp_ionsdb <- ionsdb[
        ionsdb$ion_adduct %in% adducts("positive")[,"name"],]
      tmp_nl <- nl[nl$positive == TRUE,]
      tmp_ions <- ions[ions$positive == TRUE, ]
    } else if(p == "NEG"){
      tmp_ionsdb <- ionsdb[
        ionsdb$ion_adduct %in% adducts("negative")[,"name"],]
      tmp_nl <- nl[nl$positive == FALSE,]
      tmp_ions <- ions[ions$positive == FALSE, ]
    }
    # get the mz-value of the compound "i":
    tmp <- tmp_ionsdb[tmp_ionsdb$compound_id == cmps$compound_id[i], ]
    if(p == "POS" & cmps$class[i] %in% c(
      "CAR", "LPC", "LPE", "LPS", "PC", "PE", "PS", "SM", "Cer", "HexCer", "LactCer")){
      tmp_add <- "[M+H]+"
    } else if(p == "POS" & cmps$class[i] %in% c(
      "FFA", "LPA", "LPG", "LPI", "PA", "PG", "PI", "MGDG", "DGDG", "MG", "DG", "TG")){
      tmp_add <- "[M+NH4]+"
    } else if(p == "NEG" & cmps$class[i] %in% c(
      "FFA", "LPA", "LPE", "LPG", "LPI", "LPS", "PA", "PE", "PG", "PI", "PS", "MG", "DG", "TG")){
      tmp_add <- "[M-H]-"
    } else if(p == "NEG" & cmps$class[i] %in% c(
      "CAR", "LPC", "PC", "MGDG", "DGDG", "SM", "Cer", "HexCer", "LactCer")){
      tmp_add <- "[M+CHO2]-"
    } 
    tmp <- tmp[tmp$ion_adduct %in% tmp_add, ]
    i_mz <- tmp[, "ion_mz"]
    if(cmps$study[i] == 1){
      i_spectra <- dda_spectra_tissues_POS[dda_spectra_tissues_POS$peak_id %in% "9999"]
      for(s in c("tissues", "maturation")){
        dda_xdata <- get(paste("dda_xdata", s, p, sep = "_"))
        dda_spectra <- get(paste("dda_spectra", s, p, sep = "_"))
        # get the feature ID of the compound "i":
        i_ft <- pks$FT[
          # select pks from study "s" and polarity "p":
          pks$dataset == paste(s, p, sep = "_") & 
            # pks from compound "i"
            pks$target_compound_id == cmps$compound_id[i] & 
            # pks from adduct "tmp_add"
            pks$target_ion_adduct == tmp_add]
        if(length(i_ft) > 0){
          ft <- get(paste("ft", s, p, sep = "_"))
          # get the peak IDs of the feature "i":
          i_id <- rownames(
            chromPeaks(dda_xdata)[ft[i_ft, "peakidx"][[1]],])
          # get the MS2 spectras of the peaks:
          is_spectra <- dda_spectra[dda_spectra$peak_id %in% i_id]
          if(p == "NEG" & cmps$compound[i] == "PA 18:1_18:1"){
            sps_2 <- addProcessing(is_spectra, max_peak)
            idx <- which(MsCoreUtils::between(unlist(mz(sps_2)), 417.241149 + 0.01 * c(-1, 1)))
            is_spectra <- is_spectra[idx]
          } else if(p == "NEG" & cmps$compound[i] == "PA 18:0_18:2"){
            i_ft <- pks$FT[pks$dataset == paste(s, p, sep = "_") & 
                             pks$target_name == "PA 36:2" & 
                             pks$target_ion_adduct == tmp_add]
            i_id <- rownames(
              chromPeaks(dda_xdata)[ft[i_ft, "peakidx"][[1]],])
            is_spectra <- dda_spectra[dda_spectra$peak_id %in% i_id]
            sps_2 <- addProcessing(is_spectra, max_peak)
            idx <- which(MsCoreUtils::between(unlist(mz(sps_2)), 419.256800 + 0.01 * c(-1, 1)))
            is_spectra <- is_spectra[idx]
          } 
          i_spectra <- c(i_spectra, is_spectra)
        } # close "if(length(i_ft) > 0)"
      } # close study "s"
      mycol <- 1
      h <- 0
    } else {
      i_spectra <- filterPrecursorMzRange(ms2, i_mz + 0.01 * c(-1, 1))
      i_spectra <- filterRt(i_spectra, cmps$RT[i]*60 + 10 * c(-1, 1))
      mycol <- 2
      h <- 0
    }
    # when for the study compound "i" there was also the standard:
    if(cmps$study[i] == 1 & cmps$std[i] == 1){
      i_std <- filterPrecursorMzRange(ms2, i_mz + 0.01 * c(-1, 1))
      i_std <- filterRt(i_std, cmps$RT[i]*60 + 10 * c(-1, 1))
      i_std <- Spectra::combineSpectra(
        i_std, intensityFun = base::sum, mzFun = base::mean, 
        tolerance = 0.01, minProp = 0.5, peaks = "intersect", 
        weighted = TRUE)
      if(length(i_std) > 0){
        i_std <- ms2_ann(mz = unlist(mz(i_std)), 
                         i = unlist(intensity(i_std)),
                         add = tmp_add, nl = tmp_nl)
        h <- -1.05
      } else {
        h <- 0
      }
      if(length(i_spectra) == 0){
        i_std <- filterPrecursorMzRange(ms2, i_mz + 0.01 * c(-1, 1))
        i_spectra <- filterRt(i_std, cmps$RT[i]*60 + 10 * c(-1, 1))
        mycol <- 2
        h <- 0
      }
    }
    if(length(i_spectra) > 0){
      # combine the MS2 spectras:
      register(SerialParam())
      ms2comb <- Spectra::combineSpectra(
        i_spectra, intensityFun = base::sum, mzFun = base::mean, 
        tolerance = 0.01, minProp = 0.5, peaks = "intersect", 
        weighted = TRUE)
      if(length(ms2comb) > 1){
        ms2comb <- Spectra::combineSpectra(
          ms2comb, intensityFun = base::sum, mzFun = base::mean, 
          tolerance = 0.01, minProp = 0.5, peaks = "intersect", 
          weighted = TRUE)
      }
      ms2comb$compound <- cmps$compound[i]
      ms2comb$adduct <- tmp_add
      sps <- ms2_ann(mz = mz(ms2comb)[[1]], 
                     i = intensity(ms2comb)[[1]],
                     add = tmp_add, nl = tmp_nl)
      ins <- matchWithPpm(sps$mz, tmp_ions$mz, ppm = 10)
      names(ins) <- seq(nrow(sps))
      ins <- ins[lapply(ins,length)>0]
      if(length(ins) > 0){
        for(j in seq(length(ins))){
          sps$ann[as.numeric(names(ins)[j])] <- tmp_ions$name[ins[[j]]]
        }
      }
      if(cmps$class[i] %in% c("mPA", "dmPA", "PE", "PG", "PI")){
        xmin <- 200
      } else {
        xmin <- min(sps$mz)
      }
      # plot the MS2:
      plot(sps$mz, sps$i, type = "h", bty = "l", col = mycol, 
           xlab = "m/z", ylab = "intensity", 
           xlim = c(xmin, precursorMz(ms2comb)), ylim = c(h, 1.15),
           main = paste("\n", 
                        sprintf("%.4f", precursorMz(ms2comb)), "@",
                        sprintf("%.2f", rtime(ms2comb)/60)))
      idx <- which(sps$i > 0.2)
      ann <- as.character(sps$ann[idx])
      ann[is.na(ann)] <- ""
      text(sps$mz[idx], sps$i[idx], 
           paste(round(sps$mz[idx], 4), "\n", ann), 
           cex = 0.8, pos = 3, offset = 0, col = mycol)
      df <- DataFrame(
        msLevel = 2L, 
        precursorMz = precursorMz(ms2comb),
        rtime = rtime(ms2comb),
        compound_id = cmps$compound_id[i],
        instrument = "orbitrap"
      )
      if(p == "POS"){
        df$polarity <- 1L
      } else if(p == "NEG"){
        df$polarity <- 0L
      }
      df$mz <- list(c(sps$mz))
      df$intensity <- list(c(sps$i))
      df <- Spectra(df)
      cdb <- insertSpectra(cdb, spectra = df,
                           columns = c("compound_id", "msLevel",
                                       "precursorMz", "instrument"))
      # when for the study compound "i" there is also the standard:
      if(cmps$study[i] == 1 & cmps$std[i] == 1 & h < 0){
        if(length(i_std) > 0){
          points(i_std$mz, i_std$i*(-1), type = "h", col = 2)
          idx <- which(i_std$i > 0.1)
          text(i_std$mz[idx], i_std$i[idx]*(-1), 
               round(i_std$mz[idx], 4), 
               cex = 0.8, pos = 1, offset = 0, col = 2)
        }
      }
      # add the name of the compound
      title(paste(cmps$name[i], tmp_add), adj = 0, line = 2)
      rm(ms2comb, sps)
    } # close if(length(i_spectra) > 0)
    rm(tmp_add, i_spectra)
  } # close polarity "p"
} # end "i"
dev.off()
```

# Save output

```{r}
save(pks_maturation_NEG, pks_maturation_POS, 
     pks_tissues_NEG, pks_tissues_POS, file = "output/pks.RData")
```


# Session information

```{r session}
Sys.time()-startpoint
devtools::session_info()
```

