---
title: "Retention times"
author: "Mar Garcia-Aloy"
output: 
  BiocStyle::html_document:
    toc: false
    number_sections: false
    toc_float: false
---

```{r startpoint, include = FALSE}
startpoint <- Sys.time()
```

# Read data

```{r}
files = list.files("data/")
files = files[grep("AREA", files)]
data = list()
for(i in 1:length(files)){
  data[[i]] = read.table(paste0("data/", files[i]), 
                         sep = "\t", na.strings = "N/A", skip = 3, row.names = 1, 
                         col.names = colnames(read.table(paste0("data/", files[i]),
                                                         sep = "\t", header = T)))
  colnames(data[[i]]) = gsub(
    paste0("X", gsub("_all_AREA.txt", "", files[i]), "_"), "", 
    colnames(data[[i]]))
}
data <- do.call("rbind", data)
rm(i, files)
```


## Select only "entire" samples

```{r}
data = data[, grep("entire", colnames(data))]
```

## Delete those compounds not detected in any sample

```{r}
data <- data[!apply(
  data, 1, function(x) length(which(is.na(x)))) == ncol(data),]
```


## Replace missing data by a low-, random-value

```{r}
set.seed(20200221)
data_tf <- xcms::imputeRowMinRand(data, method = "from_to",
                                  min_fraction = 1/1000,
                                  min_fraction_from = 1/10000)
```

# PCA

```{r}
data_tf = data.frame(t(log10(data_tf)))
scaling.pareto <- BioMark::scalefun(sc.p = "pareto")
data_tf = data.frame(scaling.pareto(data_tf))

pc = prcomp(data_tf, center = F, scale. = F)

par(mfrow = c(1, 2))
plot(pc$x[,1], pc$x[,2], type = "n",
     xlab = paste0("PC1: ",
                   format(summary(pc)$importance[2,1] * 100, digits = 3), 
                   "%"),
     ylab = paste0("PC2: ",
                   format(summary(pc)$importance[2,2] * 100, digits = 3), 
                   "%"),
     xlim = c(min(pc$x[,1]) - 5, max(pc$x[,1]) + 5),
     main = "Score plot")
text(pc$x[,1], pc$x[,2], rownames(pc$x))
grid()
abline(v = 0, h = 0, lty = 2)

plot(pc$rotation[,1], pc$rotation[,2],
     xlab = "PC1", ylab = "PC2", main = "Loading plot")
grid()
abline(v = 0, h = 0, lty = 2)
```

The 1st PC separates the oxidated samples (on the left) from the non-oxidated ones (on the right),
whereas the 2nd PC separates the samples from time 1 (upper part) from the ones from time 11 (bottom part).  

There is a group of compounds that seems characteristic from non-oxidated samples, let's see which are:

```{r}
plot(pc$rotation[,1], pc$rotation[,2],
     xlab = "PC1", ylab = "PC2", main = "Loading plot")
idx = pc$rotation[,1] > 0.12
points(pc$rotation[idx,1], pc$rotation[idx,2], pch = 16, col = 2)
grid()
abline(v = 0, h = 0, lty = 2)

rownames(pc$rotation)[idx]

data[is.na(data)] = 0

for(i in 1:length(which(idx))){
  plot(t(data[which(idx)[i],]), 
       main = rownames(data)[which(idx)[i]],
       xlab = "sample", ylab = "intensity", xaxt="n")
  axis(side = 1, at = seq(ncol(data)), labels = colnames(data))
}

gplots::heatmap.2(as.matrix(data_tf), 
                  scale = "none", density.info = "none", trace = "none", 
                  distfun = function(x) dist(x,method = 'euclidean'),
                  margins = c(5, 13))
```


# Session information

```{r session}
Sys.time()-startpoint
devtools::session_info()
```
