---
title: "Stats - Tissues"
author: "Mar Garcia-Aloy"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document:
    toc: true
    number_sections: false
    toc_float: true
---

```{r startpoint, include = FALSE}
startpoint <- Sys.time()
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)
```

In this document we first do an unsupervised exploratory analysis and 
later we classify compounds by "tissue specificity" according to their abundances in identified compounds.


# Libraries

```{r, message=FALSE, warning=FALSE, echo = TRUE}
library(pander)
library(effectsize)
library(ggplot2)
library(gridExtra)
library(pheatmap)
library(RColorBrewer)


myimputer <- function(v){
  set.seed(123)
  if (sum(is.na(v)) == 0) {
    return(v)
  } else {
    napos <- which(is.na(v))
    if(min(v, na.rm = TRUE) > 2){
      newval <- runif(length(napos), 1, min(v, na.rm = TRUE)/10)
    } else {
      newval <- runif(length(napos), 0, min(v, na.rm = TRUE)/10)
    }
    out <- v
    out[napos] <- newval
    return(out)
  }
}
```

# Colours 

```{r, echo = TRUE}
col_cmps <- c(
  "#E41A1C", "pink", "#974661", "#4A72A6", "#3E8E93", "#48A462", 
  "#5D995D", "#90CC90", "#7E6E85", "#A35390", "#D16948", "#FF7F00", "#FFB716", 
  "#FFF02D", "#E1C62F", "#B97B2A", "#B97B2A", "#ea890b", 
  "#B75F49", "#DB728C", "#EC83BA", "#C28EA9", "#999999", "grey",
  "#FFFFBF50", "tomato", "tomato", "black", "white")
names(col_cmps) <- c(
  "FFA", "FFA_O", "Lyso_PA", "Lyso_PE", "Lyso_PG", "Lyso_PI", 
  "PA", "mPA", "PC", "PE", "PG", "PI", "PS", 
  "CAR", "MAG", "CER", "CERhex", "pCERhex", 
  "DAG", "DGDG", "Lyso_PC", "MGDG", "TAG", "unk_NL325",
  "X", "others", "non_FA", "unknown", "IS")

class_col <- c("#377EB8", "#4DAF4A", "#E41A1C", "#984EA3", "grey")
names(class_col) <- c("skin", "pulp", "seeds", "entire", "QC")
```


# Data import

```{r import, echo = TRUE}
bygrgrape <- TRUE # specify "TRUE" or "FALSE"

data <- read.csv("output/data_tissues.csv", row.names = 1)
data <- data[!grepl("rw|dl", rownames(data)),]

class <- gsub("_rep.*", "", gsub("pt11_", "", rownames(data)))

cmps <- read.csv("compounds.csv")
cmps$id <- cmps$compound
cmps$id <- gsub(")", "", cmps$id)
cmps$id <- gsub("\\(", "_", cmps$id)
cmps$id <- gsub("/", "_", cmps$id)
cmps$id <- gsub(":", ".", cmps$id)
cmps$id <- gsub("^_", "X_", cmps$id)

cmps <- cmps[cmps$id %in% colnames(data), ]
cmps <- cmps[!duplicated(cmps$compound),]
cmps <- cmps[order(factor(cmps$id, levels = colnames(data))),]
all(colnames(data) == cmps$id)
```

The dataset contains information about n=`r ncol(data)` compounds in n=`r nrow(data)` samples of the following classes: `r paste0(aggregate(class, list(class), function(x){sum(class == x)})[,1], " (n=", aggregate(class, list(class), function(x){sum(class == x)})[,2], ")", collapse = "; ")`.   

The samples used in this analysis are: 

```{r}
pandoc.table(as.data.frame(rownames(data)[order(rownames(data))]), 
             style = "rmarkdown", caption = "Samples used in this analysis")
```


Identified compounds are from the following classes:

```{r}
pandoc.table(as.data.frame(table(cmps$class)), style = "rmarkdown")
```



# Exploratory analysis

```{r}
par(mfrow = c(2, 3), mar = c(0, 0, 1, 0))
for(p in c("POS", "NEG")){
  for(i in c("seeds", "pulp", "skin")){
    tmp <- data.frame(t(data[class == i, ]))
    if(p == "POS"){
      tmp <- tmp[rownames(tmp) %in% cmps$id[cmps$class %in% c("DAG", "TAG")], ]
    } else if(p == "NEG"){
      tmp <- tmp[rownames(tmp) %in% c(cmps$id[
        cmps$class %in% c("PA", "mPA", "PC", "PE", "PI", "MGDG", "DGDG")], 
        "X_Epicatechin"), ]
    }
    tmp$mean <- rowMeans(tmp)
    quanti <- aggregate(tmp$mean, list(cmps$class[cmps$id %in% rownames(tmp)]), 
                        function(x){sum(x, na.rm=T)})
    #quanti <- quanti[!quanti$Group.1 %in% c("IS"), ]
    slices <- quanti$x
    lbls <- quanti$Group.1
    pie(slices, lbls, main = i, col = col_cmps[as.character(lbls)])
  }
}

for(p in c("POS", "NEG")){
  for(i in c("seeds", "pulp", "skin")){
    tmp <- data.frame(t(data[class == i, ]))
    if(p == "POS"){
      tmp <- tmp[rownames(tmp) %in% cmps$id[cmps$class %in% c("DAG", "TAG")], ]
    } else if(p == "NEG"){
      tmp <- tmp[rownames(tmp) %in% c(cmps$id[
        cmps$class %in% c("PA", "mPA", "PC", "PE", "PI", "MGDG", "DGDG")], 
        "X_Epicatechin"), ]
    }
    tmp$mean <- rowMeans(tmp)
    quanti <- aggregate(tmp$mean, list(cmps$class[cmps$id %in% rownames(tmp)]), 
                        function(x){sum(x, na.rm=T)})
    quanti <- quanti[!quanti$Group.1 %in% c("IS", "unknown", "others", "non_FA"), ]
    slices <- quanti$x
    lbls <- quanti$Group.1
    pie(slices, lbls, main = i, col = col_cmps[as.character(lbls)])
  }
}
```

From the pie charts it can be seen that the most abundant lipids in all 3 tissues are `TAGs`, whereas the second most abundant class is most tissue-specific: it is `DAG` for `seeds` and `PC` for both `pulp` and `seeds`. On the other hand, it can also be seen that `skin` seems to be more diversity rich tissue.  

Missing values (n=`r sum(is.na(data))`) in the dataset are replaced by a random value between the range obtained by dividing by 1000 and 100 of the smallest measured concentration for the corresponding compound.  

Principal component analysis (PCA) is performed on log-transformed and Pareto-scaled abundances.

```{r pca, fig.width=10}
sum(is.na(data))
set.seed(123)
dt <- apply(data, 2, myimputer)
dt <- log10(dt)
scaling.pareto <- BioMark::scalefun(sc.p="pareto")
dt <- data.frame(scaling.pareto(dt))
pca <- prcomp(dt, center = FALSE, scale. = FALSE)

par(mfrow = c(1, 2), mar = c(4, 4, 2, 0.5))
pca_s <- data.frame(pca$x)
plot(pca_s$PC1, pca_s$PC2, 
     xlab = paste0("PC1 (", sprintf("%.1f", 
                                    summary(pca)$importance[2,1]*100), "%)"), 
     ylab = paste0("PC2 (", sprintf("%.1f", 
                                    summary(pca)$importance[2,2]*100), "%)"), 
     main = "Exploratory PCA", col = "white", bty = "l")
text(pca_s$PC1, pca_s$PC2, gsub("rep", "", gsub("pt11_", "", rownames(data))))
abline(v = 0, h = 0, lty = 2, col = "grey")

pca_l <- data.frame(pca$rotation)
plot(pca_l$PC1, pca_l$PC2, xlab = "PC1", ylab = "PC2", bty = "l")
abline(v = 0, h = 0, lty = 2, col = "grey")
#plot_ly(x = tmp$PC1, y = tmp$PC2, text = colnames(data))

plot(pca_s$PC1, pca_s$PC2, 
     xlab = paste0("PC1 (", sprintf("%.1f", summary(pca)$importance[2,1]*100), "%)"), 
     ylab = paste0("PC2 (", sprintf("%.1f", summary(pca)$importance[2,2]*100), "%)"), 
     main = "Exploratory PCA", col = class_col[class], bty = "l", pch = 16)
text(pca_s$PC1, pca_s$PC2, 
     gsub("rep", "", gsub("pt11_", "", rownames(data))), pos = 1, cex = 0.6)
abline(v = 0, h = 0, lty = 2, col = "grey")

plot(pca_l$PC1, pca_l$PC2, xlab = "PC1", ylab = "PC2", bty = "l",
     col = col_cmps[cmps$class], pch = 16)
abline(v = 0, h = 0, lty = 2, col = "grey")
text(pca_l$PC1, pca_l$PC2, cmps$compound, pos = 1, cex = 0.6)
```

Injections from the same biological/technical replicates cluster together.   
The PCA shows a separation of `seed` samples from `skin` and `pulp` samples along PC1 (explaining `r round(summary(pca)$importance[2,1]*100, 1)`% of the total variance), whereas the PC2 (`r round(summary(pca)$importance[2,2]*100, 1)` % of total variance) separates `skin` samples from `pulp` ones. 

```{r pca2, fig.height=7, fig.width=20}
par(mfrow = c(1, 2))

biplot(pca)

plot(pca_s$PC1, pca_s$PC2, 
     xlab = paste0("PC1 (", sprintf("%.1f", summary(pca)$importance[2,1]*100), "%)"), 
     ylab = paste0("PC2 (", sprintf("%.1f", summary(pca)$importance[2,2]*100), "%)"), 
     col = "white", bty = "l")
points(pca_s$PC1, pca_s$PC2, pch = 19, cex = 1.5, col = class_col[class])
text(pca_s$PC1, pca_s$PC2, gsub("rep", "", gsub("pt11_", "", rownames(data))), 
     pos = 1, cex = 0.6)
text(pca_l$PC1*min(range(pca_s$PC1) / range(pca_l$PC1)), 
     pca_l$PC2*min(range(pca_s$PC2) / range(pca_l$PC2)), 
     cmps$compound, cex = 0.5, col = col_cmps[cmps$class])
grid()
abline(v = 0, h = 0, lty = 2, col = "grey", lwd = 2)
```

`r if(!bygrgrape){"Biplot shows that it seems that, as expected, *pulp* samples are the ones with less abundance of compounds. This is expected since is the tissue with a higher proportion of water, which could act as a 'dilution' factor and samples are expressed as gr/tissue."}` 

`r if(!bygrgrape){"Biplot also suggest, also as expected, that *TAGs* seem to be characteristic of *seed* samples and *FFA* of *skin* samples."}`

`r if(bygrgrape){"Biplot suggest that *FFAs* seem to be characteristic of *skin* samples."}`


The subsequent analyses will be applied only to samples of the classes `seeds`, `pulp` and `skin`. Therefore, below the samples from the classes `QC` and `entire` are excluded from the dataset:

```{r subset}
data <- data[grep("seeds|pulp|skin", rownames(data)),]
class <- gsub("_rep.*", "", gsub("pt11_", "", rownames(data)))

if(bygrgrape){
  # Transform the data according to the proportion of each tissue in 1 gr of grape
  class_prop <- c(0.02974026, 0.821298701, 0.148961039)
  names(class_prop) <- c("seeds", "pulp", "skin")
  dt <- data * class_prop[class]
  rownames(dt) <- rownames(data)
  data <- dt
}

sum(is.na(data))
set.seed(123)
dt <- apply(data, 2, myimputer)
dt <- log10(dt)
dt <- data.frame(scaling.pareto(dt))
pca <- prcomp(dt, center = FALSE, scale. = FALSE)

par(mfrow = c(1, 2), mar = c(4, 4, 2, 0.5))
pca_s <- data.frame(pca$x)
plot(pca_s$PC1, pca_s$PC2, 
     xlab = paste0("PC1 (", sprintf("%.1f", summary(pca)$importance[2,1]*100), "%)"), 
     ylab = paste0("PC2 (", sprintf("%.1f", summary(pca)$importance[2,2]*100), "%)"), 
     main = "Exploratory PCA", col = class_col[class], bty = "l", pch = 16)
text(pca_s$PC1, pca_s$PC2, 
     gsub("rep", "", gsub("pt11_", "", rownames(data))), pos = 1, cex = 0.6)
abline(v = 0, h = 0, lty = 2, col = "grey")

plot(pca_l$PC1, pca_l$PC2, xlab = "PC1", ylab = "PC2", bty = "l",
     col = col_cmps[cmps$class], pch = 16)
abline(v = 0, h = 0, lty = 2, col = "grey")
text(pca_l$PC1, pca_l$PC2, cmps$compound, pos = 1, cex = 0.6)

par(mfrow = c(1, 2))

biplot(pca)

plot(pca_s$PC1, pca_s$PC2, 
     xlab = paste0("PC1 (", sprintf("%.1f", 
                                    summary(pca)$importance[2,1]*100), "%)"), 
     ylab = paste0("PC2 (", sprintf("%.1f", 
                                    summary(pca)$importance[2,2]*100), "%)"), 
     col = "white", bty = "l")
points(pca_s$PC1, pca_s$PC2, pch = 19, cex = 1.5, col = class_col[class])
text(pca_s$PC1, pca_s$PC2, gsub("rep", "", gsub("pt11_", "", rownames(data))), 
     pos = 1, cex = 0.6)
text(pca_l$PC1*min(range(pca_s$PC1) / range(pca_l$PC1)), 
     pca_l$PC2*min(range(pca_s$PC2) / range(pca_l$PC2)), 
     cmps$compound, cex = 0.5, col = col_cmps[cmps$class])
grid()
abline(v = 0, h = 0, lty = 2, col = "grey", lwd = 2)
```



# Differential abundance analysis: Hedges'g

Pairwise comparisons are estimated by calculating the Hedges'g d effect size (we use it instead of Cohen's d due to the small sample size), considering as benchmark value an absolute value > 2 to select relevant differences among tissues.   
Effect size tells you how much one group differs from another.    
A g of 1 indicates the two groups differ by 1 standard deviation, a g of 2 indicates they differ by 2 standard deviations, and so on. 

```{r}
set.seed(123)
dt <- apply(data, 2, myimputer)
dt <- log10(dt)
class_pair <- combn(c("seeds", "pulp", "skin"), 2)
for(i in seq(ncol(class_pair))){
  a <- class_pair[1,i]
  b <- class_pair[2,i]
  dt_a <- dt[class == a, ]
  dt_b <- dt[class == b,]
  hg <- c()
  for(j in seq(ncol(dt))){
    hg <- c(hg, (hedges_g(dt_a[,j], dt_b[,j]))$Hedges_g)
  }
  cmps <- cbind(cmps, hg)
  colnames(cmps)[ncol(cmps)] <- paste("hedge", a, b, sep = "_")
}
hist(abs(c(cmps$hedge_seeds_pulp, cmps$hedge_seeds_skin, cmps$hedge_pulp_skin)), 
     main = "hist of abs(Hedge'g) > 2", xlab = "abs(Hedge'g)")
```

Out of `r 3*ncol(data)` pairwise comparisons done (3x`r ncol(data)` compounds), `r sum(abs(c(cmps$hedge_seeds_pulp, cmps$hedge_seeds_skin, cmps$hedge_pulp_skin)) > 2)` resulted with a g value >2. They correspond to a total of n=`r sum(apply(cmps[,grep("hedge_", colnames(cmps))], 1, function(x){max(abs(x), na.rm = T)}) > 2)` compounds (`r round((sum(apply(cmps[,grep("hedge_", colnames(cmps))], 1, function(x){max(abs(x), na.rm = T)}) > 2)/ncol(data))*100)`%).  

The relationships between the different batches for each compound are then automatically established:

```{r, fig.width=10}
beh <- c()
for(i in seq(ncol(dt))){
  x <- sum(abs(c(cmps$hedge_seeds_pulp[i], cmps$hedge_seeds_skin[i], 
                 cmps$hedge_pulp_skin[i])) > 2)
  tmp <- data.frame(aggregate(data[,i], by = list(class), function(x){
    mean(x, na.rm = T)}))
  tmp[is.na(tmp)] <- 0
  high <- tmp$Group.1[which.max(tmp$x)]
  low <- tmp$Group.1[which.min(tmp$x)]
  med <- unique(class)[!unique(class) %in% c(high, low)]
  if(x == 3){
    beh <- c(beh, paste(high, med, low, sep = " > "))
  } else if(x == 2) {
    tmp <- gsub("hedge_", "", names(cmps[i,(ncol(cmps)-2):ncol(cmps)])[which(
      abs(cmps[i,(ncol(cmps)-2):ncol(cmps)]) < 2)])
    if(high %in% unlist(strsplit(tmp, "_"))){
      beh <- c(beh, paste0("(", gsub("_", " & ", tmp), ") > ", low))
    } else if(low %in% unlist(strsplit(tmp, "_"))){
      beh <- c(beh, paste0(high, " > (", gsub("_", " & ", tmp), ")"))
    } else(print(i))
  } else if (x == 1){
    tmp <- gsub("hedge_", "", names(cmps[i,(ncol(cmps)-2):ncol(cmps)])[which(
      abs(cmps[i,(ncol(cmps)-2):ncol(cmps)]) < 2)])
    if(high %in% unlist(strsplit(tmp, "_")) &
       low %in% unlist(strsplit(tmp, "_"))){
      beh <- c(beh, paste(high, low, sep = " > "))
    } else {print(i)}
  } else {
    beh <- c(beh, "ND")
  }
}
names(beh) <- cmps$compound
beh <- factor(beh, levels = c(
  "seeds > (pulp & skin)", "seeds > pulp", "seeds > pulp > skin", 
  "seeds > skin", "seeds > skin > pulp",
  "(pulp & skin) > seeds",
  "pulp > (seeds & skin)", "pulp > seeds", "pulp > seeds > skin", 
  "pulp > skin", "pulp > skin > seeds",
  "(seeds & skin) > pulp",
  "skin > (seeds & pulp)", "skin > pulp", "skin > pulp > seeds", 
  "skin > seeds", "skin > seeds > pulp",
  "(seeds & pulp) > skin",
  "ND"
))
cmps$beh <- beh
table(beh)

par(mfrow = c(1, 2), mar = c(4, 4, 2, 0.5))
plot(pca_s$PC1, pca_s$PC2, 
     xlab = paste0("PC1 (", sprintf("%.1f", 
                                    summary(pca)$importance[2,1]*100), "%)"), 
     ylab = paste0("PC2 (", sprintf("%.1f", 
                                    summary(pca)$importance[2,2]*100), "%)"), 
     bty = "l", pch = 19, col = class_col[class])
abline(v = 0, h = 0, lty = 2, col = "grey")

plot(pca_l$PC1, pca_l$PC2, xlab = "PC1", ylab = "PC2", bty = "l")
abline(v = 0, h = 0, lty = 2, col = "grey")

idx <- which(beh %in% c(
  "seeds > (pulp & skin)", "seeds > pulp > skin", "seeds > pulp", 
  "seeds > skin", "seeds > skin > pulp"))
points(pca_l$PC1[idx], pca_l$PC2[idx], col = "#377EB8", pch = 16)

idx <- which(beh %in% c(
  "skin > (seeds & pulp)", "skin > pulp", "skin > pulp > seeds", 
  "skin > seeds", "skin > seeds > pulp"
))
points(pca_l$PC1[idx], pca_l$PC2[idx], col = "#4DAF4A", pch = 16)

idx <- which(beh %in% c(
  "pulp > (seeds & skin)", "pulp > seeds > skin", "pulp > skin > seeds", 
  "pulp > seeds", "pulp > skin"
))
points(pca_l$PC1[idx], pca_l$PC2[idx], col = "#E41A1C", pch = 16)

idx <- which(beh == "(pulp & skin) > seeds")
points(pca_l$PC1[idx], pca_l$PC2[idx], col = "#A6CEE3", pch = 16)

idx <- which(beh == "(seeds & skin) > pulp")
points(pca_l$PC1[idx], pca_l$PC2[idx], col = "#FB9A99", pch = 16)

idx <- which(beh == "(seeds & pulp) > skin")
points(pca_l$PC1[idx], pca_l$PC2[idx], col = "#B2DF8A", pch = 16)
```

Dark colors in the loading plot indicate compounds associated with the corresponding tissue, whereas light colors indicate compounds for which the corresponding tissue is the one showing the lowest abundances.  

As almost all compounds are characteristic of one (or two) particular tissue(s), we will now assess in general terms whether there is a class of compounds that is specific to a certain tissue.  
For that we check the distribution of tissue-source according to compound class.

```{r}
beh <- beh[order(beh)]
p <- list()
for(i in seq(length(beh))){
  dt <- data.frame(
    value = data[,cmps$compound == names(beh)[i]],
    class = class
  )
  dt <- dt[order(dt$class),]
  dt$na <- is.na(dt$value)
  dt$value[is.na(dt$value)] <- 1
  dt$class_n <- as.numeric(factor(dt$class)) + rep(seq(-0.1, 0.1, by = 0.07), 3)
  p[[i]] <- ggplot(data = dt, aes(x = class_n, y = value)) + 
    geom_point(aes(color = class, pch = na), size = 2) +
    #geom_segment(aes(x=class_n, xend=class_n, y=0, yend=value, color = class)) +
    ggtitle(paste(names(beh)[i]), beh[i]) +
    xlab("") + ylab("") + theme_bw()+ theme(legend.position = "none") +
    #scale_color_manual(values=c("#FFD92F", "#E41A1C", "#4DAF4A")) + 
    scale_y_log10() + 
    #scale_y_log10(limits = c(1, NA)) + 
    scale_shape_manual(values = c(19,1)) +
    scale_x_continuous(breaks = c(1,2,3), labels = c("pulp", "seeds", "skin")) 
}
if(bygrgrape){
  fl <- "output/tissues_boxplots_gr_grape.pdf"
} else{
  fl <- "output/tissues_boxplots_gr_tissue.pdf"
}
ggsave(
  filename = fl,
  plot = marrangeGrob(p, nrow = 4, ncol = 4),
  width = 15, height = 12
)

cmps$beh2 <- "ND"
cmps$beh2[cmps$beh %in% c(
  "seeds > (pulp & skin)", "seeds > pulp", "seeds > pulp > skin", "seeds > skin", 
  "seeds > skin > pulp")] <- "seeds"
cmps$beh2[cmps$beh %in% c(
  "pulp > (seeds & skin)", "pulp > seeds", "pulp > seeds > skin", "pulp > skin", 
  "pulp > skin > seeds")] <- "pulp"
cmps$beh2[cmps$beh %in% c(
  "skin > (seeds & pulp)", "skin > pulp", "skin > pulp > seeds", 
  "skin > seeds", "skin > seeds > pulp")] <- "skin"
cmps$beh2[cmps$beh == "(pulp & skin) > seeds"] <- "pulp & skin"
cmps$beh2[cmps$beh == "(seeds & skin) > pulp"] <- "seeds & skin"
cmps$beh2[cmps$beh == "(seeds & pulp) > skin"] <- "seeds & pulp"

cmps$beh2 <- factor(cmps$beh2, levels = c(
  "seeds", "pulp", "skin", "seeds & pulp", "seeds & skin", "pulp & skin", "ND"
))
pandoc.table(table(cmps$class, cmps$beh2), 
             style = "rmarkdown", split.table = Inf)
tmp2 <- prop.table(table(cmps$class, cmps$beh2), 1)*100
pandoc.table(round(tmp2), style = "rmarkdown", split.table = Inf)
```

The following heatmap shows the distribution of tissue-sources by compound class.  
Proportions of compounds in different tissues are depicted with a color scale where dark blue represents the categories with higher proportions and light blue / white the lowest ones.

```{r}
pheatmap(tmp2, color = colorRampPalette(brewer.pal(n = 9, name ="Blues"))(100), 
         cutree_cols = 4, cutree_rows = 4)
pheatmap(tmp2[,c("seeds", "pulp", "skin")], 
         color = colorRampPalette(brewer.pal(n = 9, name ="Blues"))(100), 
         cutree_cols = 3, cutree_rows = 3)
```

`r if(!bygrgrape){"Heatmap clearly shows that all *FFA* identified are characteristic of *skin* samples, whereas most of *PA* and half of *pCERhex* compounds are also particular of this tissue, when evaluating the abudance detected by gr/tissue."}`

`r if(!bygrgrape){"On the other hand, all identified *FFA_O* and most of *DAG*, *TAG* and, to lesser extend, *PE* are particular of *seed* samples. Also the only *Lyso_PC* indentified presented the highest abundances in this type of samples."}`  

`r if(!bygrgrape){"Finally, it also confirms that *pulp* samples seems that are not particular of any compound. But here it is also necessary to keep in mind that these analyses are done by gr/tissue and it is a tissue with much higher proportions of water, which acts as a 'dilution' factor."}`

`r if(bygrgrape){"Heatmap clearly shows that all *FFA* and most *mPA* are characteristic of *skin* samples."}`

`r if(bygrgrape){"It also shows that *pulp* is the tissue that provides the highest amount of most of identified compounds, probably because it is the tissues high a highest amount by a grape berry (it is important to remember that this data is corrected by the proportion of each tissue in a grape berry). This situation is precissely mostly the oposite when evaluating the abundances of compounds by gr/tissue instead of by gr/grape, because in that case the pulp is the tissue with the highest amount of water, which probably is acting as a diluting factor. Specifically the output of the heatmap shows that all *CERhex*, *PA*, *PE* and *PI. Also the only *Lyso_PC* and *PS* indentified presented the highest abundances in this type of samples. Also most of *pCERhex*, *PC*, *PG*, *MGDG*, *DGDG* and *TAG* are more provided with higher amounts in *pulp*."}`

`r if(bygrgrape){"Finally, regarding *seeds* it seems that they are providing the highest amounts of *DAG*, alghouth for few of them the tissue with the higesth proportions is *pulp*."}`

```{r new-approach}
for(i in seq(length(unique(class)))){
  tmp <- colMeans(data[class == unique(class)[i],], na.rm = T)
  cmps <- cbind(cmps, tmp)
  colnames(cmps)[ncol(cmps)] <- unique(class)[i]
  cmps <- cbind(cmps, as.integer(cut(tmp, quantile(tmp, probs = 0:4/4, na.rm = T),
                                     include.lowest=TRUE)))
  colnames(cmps)[ncol(cmps)] <- paste(unique(class)[i], "q", sep = "_")
  cmps[is.na(cmps[,ncol(cmps)]),ncol(cmps)] <- 1
}
cmps$sd_q <- apply(cmps[,c("seeds_q", "pulp_q", "skin_q")], 1, sd)

set.seed(123)
dt <- apply(data, 2, myimputer)
dt <- log10(dt)
cmps$r2 <- NA
for(i in seq(ncol(dt))){
  cmps$r2[i] <- summary(lm(dt[,i]~class))$r.squared
}
```

# ANOVA

```{r}
all(colnames(data) == cmps$id)
set.seed(123)
dt <- apply(data, 2, myimputer)
dt <- log10(dt)
cmps$aov <- NA
cmps$seeds_pulp <- NA
cmps$skin_pulp <- NA
cmps$skin_seeds <- NA
for(i in seq(nrow(cmps))){
  cmps$aov[i] <- summary(aov(dt[,i]~class))[[1]][1,"Pr(>F)"]
  cmps$seeds_pulp[i] <- TukeyHSD(aov(dt[,i]~class))$class["seeds-pulp","p adj"]
  cmps$skin_pulp[i] <- TukeyHSD(aov(dt[,i]~class))$class["skin-pulp","p adj"]
  cmps$skin_seeds[i] <- TukeyHSD(aov(dt[,i]~class))$class["skin-seeds","p adj"]
}
cmps$padj <- p.adjust(cmps$aov, "BH")
nrow(cmps)
sum(cmps$padj < 0.05 & apply(cmps[, c("hedge_seeds_pulp", "hedge_seeds_skin", "hedge_pulp_skin")], 1, function(x) max(abs(x))) > 2)

rownames(dt) <- gsub("rep", "", gsub("pt11_", "", rownames(dt)))
dt_class <- data.frame(class = class)
rownames(dt_class) <- rownames(dt)
dt_cmp <- data.frame(cmp = cmps$class)
rownames(dt_cmp) <- colnames(dt)
mycols <- list(class = class_col,
                 cmp = col_cmps)
pheatmap(data.frame(scaling.pareto(dt)),
         annotation_row = dt_class, cutree_rows = 3, 
         annotation_colors = mycols,
         annotation_col = dt_cmp, cutree_cols = 3)
```


# Session information

```{r session}
Sys.time()-startpoint
devtools::session_info()
```


