---
title: "RT and ppm deviations"
output: html_document
---

The scope of this document is to check which are the RT and m/z deviations of each experiment.  
For that I´ll use the following types of samples: `solvent` (we shouldn't see "anything"), `blank` (we should see "only" the IS) and `STDmix` (we should see the ions related with the 77 compounds included in this mix). There are 2 injections of each type of sample. Therefore, we will use 6 samples in total.  

The workflow will be the following:  

  - Data processing with XCMS  
  - Feature grouping  
  - Identification based on theoretical database of IS and STDmix compounds  
  
  
# Preliminaries

```{r preliminaries, message=FALSE, warning=FALSE}
polarity <- "NEG" # specify "POS" or "NEG"
study <- "tissues" # specify "maturation" or "tissues"

if(polarity == "NEG"){
  int <- 1e5
} else if(polarity == "POS"){
  int <- 1e6
}

thr_int <- 5e7


library(xcms)
library(DT)
library(plotly)

startpoint <- Sys.time()
```

# XCMS processing

## Data import

```{r data-import}
#  Get filenames
injections <- list.files(paste0(study, "/data/", polarity, "_FS_fixed/"))
injections <- data.frame(
  filename = injections[grepl("blank|STDmix|solv_rep1|solv_rep2", injections)])
tmp <- strsplit(gsub(".mzData", "", injections$filename), "_")
tmp <- data.frame(do.call(rbind, tmp))
tmp$X2 <- paste(tmp$X2, tmp$X3, sep = "_")
tmp <- tmp[,-3]
colnames(tmp) <- c("order", "project", "pt", "tissue", "rep", "polarity", "mode")
injections <- cbind(injections, tmp)
rm(tmp)

# Read the data
data_raw <- readMSData(
  paste0(study, "/data/", polarity, "_FS_fixed/" , injections$filename),
  pdata = as(AnnotatedDataFrame(injections),
             "NAnnotatedDataFrame"), 
  mode = "onDisk")
```


## Peak detection

```{r peak-detection, message=FALSE}
cwp <- CentWaveParam(ppm = 20,
                     peakwidth = c(2, 20),
                     prefilter = c(5, int),
                     snthresh = 5,
                     noise = 1000,
                     mzdiff = 0.001,
                     integrate = 2)
xdata <- findChromPeaks(data_raw, param = cwp)
```

### Filter low-intensity peaks

```{r filter-low-intensity-peaks}
xdata <- refineChromPeaks(xdata, 
                          param = FilterIntensityParam(
                            nValues = 5, threshold = int)
)
```

### Peak post-processing

```{r peak-post-processing}
mnp <- MergeNeighboringPeaksParam(
  expandRt = 2, 
  expandMz = 0.001, 
  ppm = 10,
  minProp = 0.66)
register(SerialParam())
xdata <- refineChromPeaks(xdata, param = mnp)
```


## Aligment

```{r aligment}
# Define the hook peaks
pdp <- PeakDensityParam(sampleGroups = xdata$tissue,
                        minFraction = 1,
                        binSize = 0.02,
                        bw = 3)
xdata <- groupChromPeaks(xdata, param = pdp)

# Perform the aligment
pgp <- PeakGroupsParam(span = 0.3,
                       subset = which(xdata$tissue == "STDmix"),
                       subsetAdjust = "average",
                       minFraction = 1)
xdata <- adjustRtime(xdata, param = pgp)
```

## Correspondance

```{r correspondance}
pdp <- PeakDensityParam(sampleGroups = xdata$tissue,
                        minFraction = 1,
                        minSamples = 2,
                        binSize = 0.02,
                        bw = 1)
xdata <- groupChromPeaks(xdata, param = pdp)
```

## Peak filling

```{r peak-filling}
fcp <- ChromPeakAreaParam()
xdata <- fillChromPeaks(xdata, param = fcp)

rm(cwp, mnp, pgp, pdp, fcp, int, injections, data_raw)
```


# Feature grouping

```{r feature-grouping, message=FALSE}
source("../../R/ft_group.R")

datax <- dt_preparation(xobj = xdata, class = "tissue", 
                        mznoise = c(158.975, 184.0731, 219.174, 221.154, 
                                    251.164, 267.196, 269.2115, 283.699, 
                                    415.284))

features <- datax[["features"]]
features <- features[features$npeaks == 2 & features$blank == 0 &                       
                       features$solv == 0 & features$STDmix == 2, ]
features <- features[features$prop_solv > thr_int |
                       features$prop_blank > thr_int |
                       features$prop_STDmix > thr_int, ]
datax[["features"]] <- features

load(paste0(study, "/data/RData/MS2_library_", polarity, ".RData"))
datax <- ft_grouping(datax = datax, 
                     MS2x = ms2list)

features <- isotopologues(features = datax[["features"]])

features <- fg_grouping(data = datax[["data"]],
                        features = features)
```


# Identification

```{r identification}
cmps <- read.csv("compounds_RT_formula.csv")
IS <- cmps[cmps$class == "IS", ]
old <- read.csv("z_old/compounds.csv")
old <- old[old$type == "MIXnativi",]
mix <- cmps[cmps$ID %in% old$name, ]
mix2 <- old[!old$name %in% cmps$ID, ]
mix2 <- subset(mix2, select = c("name", "RT", "class", "formula", "C"))
colnames(mix2)[1] <- "ID"
mix2$POS <- NA
mix2$NEG <- NA
mix2$RT <- mix2$RT/60
cmps <- rbind(IS, mix, mix2)
cmps$RT <- cmps$RT*60
cmps$mass <- NA
for(i in seq(nrow(cmps))){
  cmps$mass[i] <- getMolecule(cmps$formula[i])$exactmass
}
rm(IS, old, mix, mix2, i)

ann <- c(
  "[M+H]+", "13C[M+H]+", "(2)13C[M+H]+", "[M]+",
  "[M+NH4]+", "13C[M+NH4]+", "(2)13C[M+NH4]+", "[M]+ +NH3",
  "[M+Na]+", "13C[M+Na]+", "[M+K]+", 
  "[M+C2H8N]+", "13C[M+C2H8N]+",
  "[2M+H]+", "13C[2M+H]+", "(2)13C[2M+H]+", "[3M+H]+", "13C[3M+H]+",
  "[2M+Na]+", "13C[2M+Na]+",
  
  "[M-H]-", "13C[M-H]-", "(2)13C[M-H]-", "(3)13C[M-H]-",
  "[M-H+HCOOH]-", "13C[M-H+HCOOH]-", 
  "[2M-H]-", "13C[2M-H]-", "(2)13C[2M-H]-", "(3)13C[2M-H]-",
  "[2M-2H+Na]-",
  "[3M-H]-"
)


dt_ids <- identification(features, cmps, MS2x = ms2list)

rm(datax, features, ms2list, ann, thr_int)



dt_fg <- dt_ids[["dt_fg"]]
features <- dt_ids[["features"]]
features$d_RT <- NA
features$ppm <- NA
for(i in seq(nrow(dt_fg))){
  idx <- which(cmps$ID %in% unlist(strsplit(dt_fg$compound[i], "; ")))
  if(length(idx) == 1){
    i.features <- features[features$FGx == dt_fg$FG[i], ]
    for(j in seq(nrow(i.features))){
      if(grepl("13C", i.features$assignation[j])){
        mz.thr <- unlist(mass2mz(
          getMolecule(cmps$formula[idx])$exactmass, 
          gsub("13C", "", i.features$assignation[j]))) + 1.003355
      } else {
        mz.thr <- unlist(mass2mz(getMolecule(cmps$formula[idx])$exactmass, 
                                 i.features$assignation[j]))
      } # close "if(grepl("13C", i.features$assignation[j]))"
      
      i.features$ppm[j] <- ppm_dev(i.features$mzmed[j], mz.thr)
      
    } # close i.features "j"
    
    i.features$d_RT <- i.features$rtmed - cmps$RT[idx]
    
    features[rownames(i.features), "ppm"] <- i.features[, "ppm"]
    features[rownames(i.features), "d_RT"] <- i.features[, "d_RT"]
    
  } # close "if(length(idx) > 0)"
} # close dt_fg "i"
rm(i, idx, i.features, j, mz.thr)

cmp_id <- data.frame(table(dt_fg$color))
rownames(cmp_id) <- c("unknown", "identified")
cmp_id$perc <-  round((cmp_id$Freq/sum(cmp_id$Freq))*100, 1)

ft_id <- data.frame(table(!is.na(features$d_RT)))
rownames(ft_id) <- c("unknown", "identified")
ft_id$perc <-  round((ft_id$Freq/sum(ft_id$Freq))*100, 1)
```

## Tables

```{r id-table}
datatable(dt_fg[, c(1:5,9)], rownames = FALSE,  
          options = list(dom = 't', pageLength = nrow(dt_fg))) %>% 
  formatStyle("color", target = "row", 
              backgroundColor = styleEqual(
                c(0, 1, 2, 3), c('', 'lightblue', '#C0819E60', 'lightgreen')))

datatable(t(cmp_id)[-1,], caption = "Compounds",  
          options = list(dom = 't', autoWidth = TRUE))
datatable(t(ft_id)[-1,], caption = "Features",  
          options = list(dom = 't'))
```

## Plots

```{r}
features <- features[!is.na(features$ppm), ]
plot_ly(x = features$d_RT, y = features$ppm, 
                text = paste(features$FGx, features$assignation), 
                color = features$FGx, colors = "Set2") %>%
  layout(xaxis = list(title = "d(RT)"), yaxis = list(title = "ppm"))
plot_ly(x = (features$rtmed - features$d_RT), y = features$d_RT,
        text = paste(features$FGx, features$assignation), 
                color = features$FGx, colors = "Set2") %>%
  layout(xaxis = list(title = "RT - theoretical"), 
         yaxis = list(title = "d(RT)"))
tapply(features$ppm, features$polarity, summary)
tapply(features$d_RT, features$polarity, summary)
```


# Deviations

```{r}
dt <- data.frame(
  ppm = c(sprintf("%.1f", min(features$ppm)), sprintf("%.1f", max(features$ppm))),
  RT = c(round(min(features$d_RT)), round(max(features$d_RT)))
)
rownames(dt) <- c("min", "max")
datatable(dt, options = list(dom = 't'))
```


# Session information

```{r session}
Sys.time()-startpoint
devtools::session_info()
```